---
title: "A Continuum Mechanics Perspective"
---

Consider a body $\mathcal{B}$ in three-dimensional Euclidean space $\mathbb{E}^3$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 4
\begin{tikzpicture}

    % Draw the body as a continuum potato
    \draw[fill=gray!30] plot[smooth cycle, tension=0.7] coordinates {
        (3.0, 0.3) (2.6, 1.6) (1.2, 2.5) (-0.5, 2.3) (-2.2, 1.8)
        (-3.3, 0.4) (-3.0, -0.9) (-1.5, -2.1) (0.7, -2.0) (2.4, -1.1)
    };

    % Draw applied forces as arrows
    \draw[->, thick] (-2.0, 1.6) -- (-4.0, 2.4) node[anchor=south east] {$\mathbf{F}_1$};
    \draw[->, thick] (2.5, -0.9) -- (4.3, -2.0) node[anchor=north west] {$\mathbf{F}_2$};

    % Draw a force distribution along the top surface
    \foreach \x/\y in {-0.5/2.3, 0.3/2.5, 1.1/2.5, 1.8/2.2, 2.4/1.6} {
        \draw[->, thick] (\x, \y) -- (\x+0.25, \y+0.55);
    }

    % Label the force distribution
    \node[anchor=south] at (0.8, 3.0) {Force distribution};

\end{tikzpicture}
```

In general, bodies are deformable continua. We depict continua as in the above figure and refer to them as *continuum potatoes*.

Point or distributed forces or moments can be applied to a body's surface or throughout its volume. For example, a body resting on a table has contact forces on the surface touching the table, while weight acts on all points in the body.

::: {.callout-tip title="Think!"}
**Question:** How do we know if a body has moved or deformed?

::: {.callout-note title="Answer" collapse="true"}
…
:::
:::

Consider a reference configuration for this body. A reference configuration is any configuration that can be physically occupied by the body — it could be the initial configuration, but need not be.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 6
\begin{tikzpicture}

    % Draw the reference configuration (left potato)
    \draw[fill=gray!30] plot[smooth cycle, tension=0.7] coordinates {
        (-4.3, 0.4) (-4.8, 1.6) (-6.0, 2.2) (-7.4, 1.5) (-8.1, 0.2)
        (-7.8, -1.1) (-6.2, -2.0) (-4.8, -1.3) (-4.0, -0.2)
    };

    % Draw the current configuration (right potato)
    \draw[fill=gray!10] plot[smooth cycle, tension=0.7] coordinates {
        (4.8, 0.5) (4.3, 1.9) (2.6, 2.6) (0.6, 2.1) (-0.8, 1.2)
        (-1.3, -0.2) (-0.2, -2.0) (2.0, -2.3) (3.9, -1.6) (5.3, -0.4)
    };

    % Draw the origin O
    \node[circle, fill=black, inner sep=2pt, label=left:{$O$}] at (-3,-4) {};

    % Draw the point X in the reference configuration
    \node[circle, fill=blue, inner sep=2pt, label=above:{$X$}] at (-5, 1.5) {};

    % Draw the point x in the current configuration
    \node[circle, fill=red, inner sep=2pt, label=above:{$x$}] at (3.5, 1.2) {};

    % Draw the vector R from O to X
    \draw[->, thick] (-3,-4) -- (-5, 1.5) node[midway, below left] {${\bf R}$};

    % Draw the vector r from O to x
    \draw[->, thick] (-3,-4) -- (3.5, 1.2) node[midway, below right] {${\bf r}$};

    % Label the configurations
    \node at (-6, -3) {Reference Configuration};
    \node at (2, -3) {Current Configuration};

\end{tikzpicture}
```

We denote the position vector from a chosen fixed origin to a material point in the reference configuration as ${\bf R}$. The position vector to the same material point in the current configuration is ${\bf r}$.

The motion of the body is described by
\begin{align}
    {\bf r} = \bchi({\bf R},t).
\end{align}

::: {.callout-tip title="Think!"}
**Question:** What makes a body rigid?

::: {.callout-note title="Answer" collapse="true"}
In this class, we consider only rigid bodies. If we choose two points on a rigid body, the distance between them is always constant. Also, angles between material lines on a rigid body are constant.
:::
:::

::: {.callout-tip title="Think!"}
**Question:** Under what conditions do we consider a rigid body to be a particle?

::: {.callout-note title="Answer" collapse="true"}
In the beginning of the class, we consider particles. Particles are idealized rigid bodies whose rotational inertia is ignored.
:::
:::
