---
title: "Starting from Continuum Mechanics"
---

Consider a body $\mathcal{B}$ in three-dimensional Euclidean space $\mathbb{E}^3$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}

    % Draw the "potato" as an ellipse
    \draw[fill=gray!30] (0,0) ellipse (3 and 2);

    % Draw applied forces as arrows
    \draw[->, thick] (-1.5, 1) -- (-4, 2) node[anchor=south east] {$\mathbf{F}_1$};
    \draw[->, thick] (2, -1) -- (4, -2) node[anchor=north west] {$\mathbf{F}_2$};

    % Draw a force distribution along the surface
    \foreach \angle in {40, 50, 60, 70, 80, 90, 100, 110, 120} {
        \draw[->, thick] ({3*cos(\angle)}, {2*sin(\angle)}) --
                                 ({3.5*cos(\angle)}, {2.5*sin(\angle)});
    }

    % Label the force distribution
    \node[anchor=south] at (0, 2.5) {Force distribution};

\end{tikzpicture}
```

In general, bodies are deformable continua. We depict continua as in the above figure and refer to them as *continuum potatoes*.

Point or distributed forces or moments can be applied to a body's surface or throughout its volume. For example, a body resting on a table has contact forces on the surface touching the table, while weight acts on all points in the body.

::: {.callout-tip title="Think!"}
**Question:** How do we know if a body has moved?

**Answer:** …
:::

Consider a reference configuration for this body. A reference configuration is any configuration that can be physically occupied by the body — it could be the initial configuration, but need not be.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}

    % Draw the reference configuration (left potato)
    \draw[fill=gray!30] (-6,0) ellipse (2 and 2);

    % Draw the current configuration (right potato)
    \draw[fill=gray!10] (2,0) ellipse (3 and 2);

    % Draw the origin O
    \node[circle, fill=black, inner sep=2pt, label=left:{$O$}] at (-3,-4) {};

    % Draw the point X in the reference configuration
    \node[circle, fill=blue, inner sep=2pt, label=above:{$X$}] at (-5, 1.5) {};

    % Draw the point x in the current configuration
    \node[circle, fill=red, inner sep=2pt, label=above:{$x$}] at (3.5, 1.2) {};

    % Draw the vector R from O to X
    \draw[->, thick] (-3,-4) -- (-5, 1.5) node[midway, below left] {${\bf R}$};

    % Draw the vector r from O to x
    \draw[->, thick] (-3,-4) -- (3.5, 1.2) node[midway, below right] {${\bf r}$};

    % Label the configurations
    \node at (-6, -3) {Reference Configuration};
    \node at (2, -3) {Current Configuration};

\end{tikzpicture}
```

We denote the position vector from a chosen fixed origin to a material point in the reference configuration as ${\bf R}$. The position vector to the same material point in the current configuration is ${\bf r}$.

The motion of the body is described by
\begin{align}
    {\bf r} = \bchi({\bf R},t).
\end{align}

::: {.callout-tip title="Think!"}
**Question:** What makes a body rigid?

**Answer:** In this class, we consider only rigid bodies. If we choose two points on a rigid body, the distance between them is always constant. Also, angles between material lines on a rigid body are constant.
:::

::: {.callout-tip title="Think!"}
**Question:** Under what conditions do we consider a rigid body to be a particle?

**Answer:** In the beginning of the class, we consider particles. Particles are idealized rigid bodies whose rotational inertia is ignored.
:::
