---
title: "Kinematics of Rigid Bodies"
---

In this chapter, we will denote the fixed basis by $\{{\bf E}_1,{\bf E}_2,{\bf E}_3\}$.

Consider a body $\mathcal{B}$ in motion in 3D space.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw plot[thick,smooth cycle, tension=1] coordinates {(0, 0) (2, 1) (2.5, 3.5) (1.5, 4) (0.5, 3) (-1, 2)};
    \begin{scope}[shift={(10,0)}, rotate=90, transform shape]
        \draw plot[thick,smooth cycle, tension=1] coordinates {(0, 0) (2, 1) (2.5, 3.5) (1.5, 4) (0.5, 3) (-1, 2)};
    \end{scope}
    \fill (5,-2) circle (0.1);
    \node at (5,-2.5) {$O$};
    \draw[thick,->] (5,-2) -- (2,3) node[midway,left] {${\bf X}$};
    \draw[thick,->] (5,-2) -- (7,2) node[midway,right] {${\bf x}$};
    \fill (2,3.1) circle (0.1);
    \node at (2,3.5) {$X$};
    \fill (7,2.1) circle (0.1);
    \node at (7.5,2) {$X$};
    \node at (1,4.5) {Reference configuration $\kappa_0$};
    \node at (8,3) {Current configuration $\kappa_t$};
\end{tikzpicture}
```

- A body $\mathcal{B}$ is a collection of material points (mass particles or particles).
- It is convenient to define a fixed reference configuration $\boldsymbol{\kappa}_0$ for the body: ${\bf X}=\boldsymbol{\kappa}_0$.
- The present (or current) configuration of the body is denoted by $\boldsymbol{\kappa}_t$.
- We denote a material point by $X$. The position of the material point $X$ relative to a fixed origin at the reference configuration is denoted by ${\bf X}$ and at time $t$ is denoted by ${\bf x}$.
- This motion is invertible; we can use ${\bf X}$ to uniquely define the motion as a function of $X$ and $t$: ${\bf x} = {\bchi}({\bf X},t)$.
- $\bchi$ (pronounced like "kye") is called the **motion function**. It tells you the current position vector of the material point which, at the reference configuration, had position vector ${\bf X}$.

::: {.callout-tip title="Think!"}
**Question:** What is a mathematical statement that means that a body is rigid?

::: {.callout-note title="Answer" collapse="true"}
For rigid bodies the distance between any two mass particles $X_1$ and $X_2$ remains constant for all motions:
\begin{align}
    \lnorm{\bf x}_1-{\bf x}_2\rnorm = \lnorm{\bf X}_1-{\bf X}_2\rnorm.
\end{align}
The motion preserves angles between lines on the body (unlike elastic bodies or fluids) and preserves orientation (it is not a reflection).
:::
:::

Any rigid body motion can be built in two steps:

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \begin{scope}
        \draw[fill=yellow!30] (0,0) circle(2);
        \fill (-0.7,0.7) circle(0.2);
        \fill (0.7,0.7) circle(0.2);
        \draw[line width=1pt] (-1,-0.5) arc[start angle=200, end angle=340, radius=1.2];
    \end{scope}
    \begin{scope}[shift={(5,-5)}, rotate=-45, transform shape]
        \draw[fill=yellow!30] (0,0) circle(2);
        \fill (-0.7,0.7) circle(0.2);
        \fill (0.7,0.7) circle(0.2);
        \draw[line width=1pt] (-1,-0.5) arc[start angle=200, end angle=340, radius=1.2];
    \end{scope}
    \begin{scope}[shift={(12.2,-3)}, rotate=-45, transform shape]
        \draw[fill=yellow!30] (0,0) circle(2);
        \fill (-0.7,0.7) circle(0.2);
        \fill (0.7,0.7) circle(0.2);
        \draw[line width=1pt] (-1,-0.5) arc[start angle=200, end angle=340, radius=1.2];
    \end{scope}
    \draw[thick,->] (-2,-7) -- (0,0.5) node[midway,left] {${\bf X}$};
    \node at (0,2.5) {Reference configuration};
    \draw[thick,->] (-2,-7) -- (5.5,-4.5) node[midway,below] {${\bf x}_I$};
    \node at (5,-7.5) {Intermediate configuration};
    \node at (12,-0.5) {Current configuration};
    \fill (-2,-7) circle (0.1);
    \node at (-2,-7.5) {$O$};
    \draw[thick,->] (5,-4) -- (12,-2) node[midway,above] {${\bf y}(t)$};
\end{tikzpicture}
```

1. **Rotate** the object about $O$ to the desired orientation: ${\bf x}_I = {\bf Q}(t){\bf X}$ with $\lnorm{\bf Q}(t){\bf X}\rnorm = \lnorm{\bf X}\rnorm$.
2. **Translate** via the translation vector ${\bf y}(t)$.

Then every rigid motion is:
\begin{align}
    {\bf x} = \bchi({\bf X},t) = {\bf Q}(t){\bf X}+{\bf y}(t).
\end{align}

${\bf Q}(t)$ is a linear map and is called the **rotation tensor**. Both ${\bf Q}$ and ${\bf y}$ are the same for all material points (functions of $t$ only). In this class we consider **general plane motion**, where $\bomega = \dot{\theta}{\bf E}_3$ and $\balpha = \dot{\bomega}$.

::: {.callout-tip title="Think!"}
**Question:** How do we know if a body has rotated relative to a reference configuration?

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\usetikzlibrary{arrows}
\begin{tikzpicture}
    \draw[dashed] (0,0) rectangle (2,3);
    \draw (5,0) rectangle (8,2);
    \node at (1,-1) {reference configuration};
    \node at (7,-1) {current configuration};
    \draw[red,dashed,thick] (0,0) -- (10,0);
    \draw[red,thick] (8,0) -- (8,3);
    \draw[->,>=stealth',semithick, red] (8.5,0) arc[radius=0.5, start angle=0, end angle=90];
    \node[red] at (8.6,0.6) {$\theta$};
    \draw[blue,dashed,thick] (0,0) -- (3,4.5);
    \draw[blue,thick] (2,4) -- (8,0);
    \draw[->,>=stealth',semithick, blue] (2.75,4.2) arc[radius=0.5, start angle=60, end angle=150];
    \node[blue] at (2.5,4.6) {$\theta$};
\end{tikzpicture}
```

::: {.callout-note title="Answer" collapse="true"}
We need the reference configuration to define rotation. The rotation angle $\theta$ is the angular displacement of every material line from reference to current configuration. The axis of rotation is perpendicular to the page.
:::
:::

## Corotational Bases

A *corotational basis* $\{{\bf e}_1,{\bf e}_2\}$ is a pair of vectors embedded in the material. Initially ${\bf e}_i = {\bf E}_i$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[dashed] (0,0) rectangle (2,3);
    \draw[->,thick] (0.5,0.5) -- (1.5,2) node[midway,left] {$\bf v$};
    \draw[->,thick] (0,0) -- (1,0) node[below] {${\bf e}_1$};
    \draw[->,thick] (0,0) -- (0,1) node[left] {${\bf e}_2$};
    \node at (1,-1) {reference configuration};
    \node at (7,-1) {current configuration};
    \draw[->,thick] (-2,0) -- (-1,0) node[below] {${\bf E}_1$};
    \draw[->,thick] (-2,0) -- (-2,1) node[left] {${\bf E}_2$};
    \begin{scope}[shift={(8,0)}, rotate=45, transform shape]
        \draw (0,0) rectangle (2,3);
        \draw[->,thick] (0.5,0.5) -- (1.5,2) node[midway,left] {$\bf v$};
        \draw[->,thick] (0,0) -- (1,0) node[below] {${\bf e}_1$};
        \draw[->,thick] (0,0) -- (0,1) node[left] {${\bf e}_2$};
    \end{scope}
\end{tikzpicture}
```

For a rigid body, the components of a material vector ${\bf v}$ along the corotational basis are constant throughout the motion:
\begin{align}
    {\bf v} = a{\bf e}_1+b{\bf e}_2,
\end{align}
where $a$ and $b$ are constant and ${\bf e}_1(t)$, ${\bf e}_2(t)$ evolve with the body. The corotational basis is unaffected by translation.

## Tensors

A tensor ${\bf T}$ is a *linear transformation* that transforms a vector into another vector: ${\bf T}{\bf u} = {\bf v}$. A tensor exists independent of any basis and is defined by its action on vectors.

Examples: a rotation tensor ${\bf Q}$, the identity tensor ${\bf I}$, a scaling tensor ${\bf S}$.

::: {.callout-tip title="Think!"}
**Question:** Is a rotation a linear transformation?

::: {.callout-note title="Answer" collapse="true"}
Yes. Rotating $a{\bf X}$ is the same as rotating ${\bf X}$ then scaling by $a$, and rotating ${\bf X}_1+{\bf X}_2$ equals rotating each and adding. So rotations are linear transformations — hence tensors.
:::
:::

::: {.callout-tip title="Think!"}
**Question:** Are vectors and matrices the same?

::: {.callout-note title="Answer" collapse="true"}
No. A vector is an abstract object independent of any basis; a matrix is its representation in a specific basis. Similarly, a tensor is not a matrix, but can be expressed as one in a given coordinate system.
:::
:::

## Tensor (Bun) Product

The **tensor product** ${\bf a}\otimes{\bf b}$ acts on vectors by:
\begin{align}
    \lp{\bf a}\otimes{\bf b}\rp{\bf u} = \lp{\bf b}\cdot{\bf u}\rp{\bf a}.
\end{align}
This is linear (verifiable directly), so it is a tensor.

::: {.callout-tip title="Think!"}
**Question:** What does $2{\bf E}_1\otimes{\bf E}_1+2{\bf E}_2\otimes{\bf E}_2+2{\bf E}_3\otimes{\bf E}_3$ do to a vector?

::: {.callout-note title="Answer" collapse="true"}
It scales it by a factor of two.
:::
:::

::: {.callout-tip title="Think!"}
**Question:** Show that ${\bf E}_1\otimes{\bf E}_1+{\bf E}_2\otimes{\bf E}_2+{\bf E}_3\otimes{\bf E}_3$ is the identity tensor. Show also that the rotation tensor can be expressed as ${\bf Q} = {\bf e}_x\otimes{\bf E}_x+{\bf e}_y\otimes{\bf E}_y+{\bf e}_z\otimes{\bf E}_z$.

::: {.callout-note title="Answer" collapse="true"}
Apply the definition of the tensor product term by term.
:::
:::

## Matrix Representation of Tensors

Given a basis $\{{\bf e}_1,{\bf e}_2,{\bf e}_3\}$, define $Q_{ij} \equiv \lp{\bf Q}{\bf e}_j\rp\cdot{\bf e}_i$. Then the action of ${\bf Q}$ in matrix form is:
\begin{align}
    \begin{bmatrix}x_1\\x_2\\x_3\end{bmatrix}
    =
    \begin{bmatrix}Q_{11}&Q_{12}&Q_{13}\\Q_{21}&Q_{22}&Q_{23}\\Q_{31}&Q_{32}&Q_{33}\end{bmatrix}
    \begin{bmatrix}X_1\\X_2\\X_3\end{bmatrix}.
\end{align}
The matrix representation depends on the chosen basis, but the tensor itself is basis-independent.

### Rigid Body Motion

In matrix form, rigid body motion is:
\begin{align}
    \begin{bmatrix}x_1\\x_2\\x_3\end{bmatrix}
    =
    \begin{bmatrix}Q_{11}(t)&Q_{12}(t)&Q_{13}(t)\\Q_{21}(t)&Q_{22}(t)&Q_{23}(t)\\Q_{31}(t)&Q_{32}(t)&Q_{33}(t)\end{bmatrix}
    \begin{bmatrix}X_1\\X_2\\X_3\end{bmatrix}
    +
    \begin{bmatrix}y_1\\y_2\\y_3\end{bmatrix}.
\end{align}
For $[{\bf Q}(t)]$ to be a rotation matrix: ${\bf Q}{\bf Q}^T = {\bf I}$ and $\det([{\bf Q}]) = 1$.

::: {.callout-tip title="Think!"}
**Question:** Verify that the rotation about ${\bf E}_z$ by angle $\theta(t)$ has matrix representation
\begin{align*}
    \mathtt{Q} = \begin{bmatrix}\cos\theta & -\sin\theta & 0\\\sin\theta & \cos\theta & 0\\0 & 0 & 1\end{bmatrix}
\end{align*}
using ${\bf Q} = {\bf e}_1\otimes{\bf E}_1+{\bf e}_2\otimes{\bf E}_2+{\bf e}_3\otimes{\bf E}_3$ with ${\bf e}_1 = \cos\theta\,{\bf E}_1+\sin\theta\,{\bf E}_2$ and ${\bf e}_2 = -\sin\theta\,{\bf E}_1+\cos\theta\,{\bf E}_2$.

::: {.callout-note title="Answer" collapse="true"}
Compute $Q_{ij} = \lp{\bf Q}{\bf E}_j\rp\cdot{\bf E}_i = {\bf e}_j\cdot{\bf E}_i$ to recover the matrix.
:::
:::

::: {.callout-important title="Note!"}
This matrix satisfies $\mathtt{Q}\mathtt{Q}^T = \mathtt{I}$ and $\det(\mathtt{Q})=1$, so ${\bf Q}$ is a **proper orthogonal tensor**. All proper orthogonal tensors are rotations and vice versa.
:::

## The Angular Velocity Vector

From rigid body motion, for any two material points $A$ and $B$: $({\bf x}_B-{\bf x}_A) = {\bf Q}({\bf X}_B-{\bf X}_A)$. Differentiating:
\begin{align}
    {\bf v}_B-{\bf v}_A = \dot{\bf Q}{\bf Q}^T({\bf x}_B-{\bf x}_A).
\end{align}
Since ${\bf Q}{\bf Q}^T = {\bf I}$, $\dot{\bf Q}{\bf Q}^T$ is **skew**. Every skew tensor corresponds to a vector cross product, so:
\begin{align}
    {\bf v}_B-{\bf v}_A = \bomega\times({\bf x}_B-{\bf x}_A).
\end{align}
The tensor $\bOmega = \dot{\bf Q}{\bf Q}^T$ is the **angular velocity tensor** and $\bomega$ is the **angular velocity vector**. Taking another derivative:
\begin{align}
    {\bf a}_B-{\bf a}_A = \balpha\times({\bf x}_B-{\bf x}_A)+\bomega\times(\bomega\times({\bf x}_B-{\bf x}_A))
\end{align}
where $\balpha$ is the angular acceleration.

::: {.callout-important title="Note!"}
For any corotational basis vector ${\bf e}_i$ fixed to the rigid body:
\begin{align}
    \dot{\bf e}_i = \bomega\times{\bf e}_i, \quad i=1,2,3.
\end{align}
:::

## Velocity and Acceleration of Two Material Points

Let $A$ and $B$ be material points on the rigid body. Expressing ${\bf x}_{B/A}$ on a corotational basis (constant components):
\begin{align}
    {\bf x}_{B/A} &= x{\bf e}_x+y{\bf e}_y,\\
    {\bf v}_{B/A} &= x\dot{\bf e}_x+y\dot{\bf e}_y = \bomega\times{\bf x}_{B/A},\\
    {\bf a}_{B/A} &= \balpha\times{\bf x}_{B/A}+\bomega\times(\bomega\times{\bf x}_{B/A}).
\end{align}
With $\bomega = \omega{\bf E}_z$ and $\balpha = \alpha{\bf E}_z$:
\begin{align}
    \omega\,d\omega = \alpha\,d\theta, \qquad \omega\,dt = d\theta, \qquad \alpha\,dt = d\omega.
\end{align}

## Particle in a Noninertial Frame

A noninertial frame is one that accelerates (e.g. an accelerating bus, a merry-go-round). If $A$ is fixed to the rigid body and $B$ moves relative to it:
\begin{align}
    {\bf v}_B-{\bf v}_A &= \bomega\times({\bf x}_B-{\bf x}_A)+\overset{\circ}{\bf x}_{B/A},\\
    {\bf a}_B-{\bf a}_A &= \balpha\times({\bf x}_B-{\bf x}_A)+\bomega\times\lp\bomega\times\lp{\bf x}_B-{\bf x}_A\rp\rp+2\bomega\times\overset{\circ}{\bf x}_{B/A}+\overset{\circ\circ}{\bf x}_{B/A},
\end{align}
where $\overset{\circ}{\bf x}_{B/A} = \dot{x}{\bf e}_x+\dot{y}{\bf e}_y$ and $\overset{\circ\circ}{\bf x}_{B/A} = \ddot{x}{\bf e}_x+\ddot{y}{\bf e}_y$ are the first and second **corotational rates**. The term $2\bomega\times\overset{\circ}{\bf x}_{B/A}$ is the **Coriolis acceleration** and $\bomega\times(\bomega\times{\bf x}_{B/A})$ is the **centripetal acceleration**.

::: {.callout-tip title="Think!"}
**Question:** See [this video](https://www.youtube.com/watch?v=mPsLanVS1Q8) of people playing catch on a merry-go-round. A ball is thrown on a merry-go-round rotating about fixed center $O$, with ${\bf r}_0 = -R{\bf E}_x$ and ${\bf v}_0 = v_0{\bf E}_x$. What velocity does the ball appear to have to an observer on the merry-go-round?

::: {.callout-note title="Answer" collapse="true"}
The ball is a projectile: ${\bf a} = -g{\bf E}_z$, ${\bf v} = -gt{\bf E}_z+{\bf v}_0$. From ${\bf v} = \bomega\times({\bf r}-{\bf r}_0)+\overset{\circ}{\bf x}_{P/O}$:
\begin{align}
    \overset{\circ}{\bf x}_{P/O} = -gt{\bf E}_z+v_0{\bf E}_x+\omega(v_0 t-R){\bf E}_y.
\end{align}
:::
:::

## Classifications of Rigid Body Motions

**Pure (rectilinear) translation:** $\bomega = {\bf 0}$, $\balpha = {\bf 0}$.

**Curvilinear translation:** material points follow parallel curved paths; $\bomega = {\bf 0}$, $\balpha = {\bf 0}$ (e.g. [Ferris wheel bucket](https://www.youtube.com/watch?v=LU_xVr4b2qM&t=387s)).

**Fixed point rotation:** $\bomega \neq {\bf 0}$, fixed point with ${\bf v}_O = {\bf a}_O = {\bf 0}$ (e.g. pendulum):
\begin{align}
    {\bf a}_P = \balpha\times{\bf r}_P+\bomega\times(\bomega\times{\bf r}_P).
\end{align}
The term $\balpha\times{\bf r}_P$ is the tangential acceleration; $\bomega\times(\bomega\times{\bf r}_P)$ is the centripetal acceleration.

**General motion:** combined translation and rotation.

## Instantaneous Center of Rotation

::: {.callout-tip title="Think!"}
**Question:** Draw the velocity profile of a rotating link with a fixed end.

::: {.callout-note title="Answer" collapse="true"}
```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \def\L{4}
    \def\w{2}
    \draw[thick] (0,0) -- (\L,0) node[right] {End of Link};
    \filldraw[black] (0,0) circle (2pt) node[left] {Pivot};
    \draw[->,thick] (1,0.5) arc (0:180:0.5);
    \node at (-1,1) {$\omega$};
    \foreach \r in {1,2,3,4} {
        \draw[->,red] (\r,0) -- ++(0,\r*\w*0.5) node[above] {$v = \omega r$};
    }
    \draw[->] (0,-1) -- (0,3) node[left] {$v$};
    \draw[->] (-1,0) -- (\L+1,0) node[below] {$r$};
    \draw[thick,blue] (0,0) -- (\L,\L*\w*0.5) node[right] {$v = \omega r$};
\end{tikzpicture}
```
:::
:::

A body in general plane motion is not rotating about a fixed point, but **instantaneously** it rotates about the **instantaneous center of rotation (IC)** — a point with zero velocity at that instant. If ${\bf v}_{IC} = {\bf 0}$, then for any material point $A$:
\begin{align}
    {\bf v}_A = \bomega\times({\bf r}_A-{\bf r}_{IC}).
\end{align}
So ${\bf r}_{A/IC}$ is perpendicular to ${\bf v}_A$. Knowing the directions of ${\bf v}_A$ and ${\bf v}_B$ for two material points identifies the IC.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw plot[smooth cycle, tension=1] coordinates {(0,0) (2,1) (3,2) (2.5,3.5) (1.5,4) (0.5,3) (-1,2)};
    \draw (6,2) -- (1,1) node[below] {$A$};
    \draw[->] (1,1) -- (0.7,2) node[above] {${\bf v}_A$};
    \draw (6,2) -- (2,3) node[below] {$B$};
    \draw[->] (2,3) -- (2.3,4) node[above] {${\bf v}_B$};
    \node at (6.5,2) {IC};
\end{tikzpicture}
```

For the IC:
\begin{align}
    {\bf v}_P = \bomega\times{\bf r}_{P/IC}, \qquad
    {\bf a}_P = \balpha\times{\bf r}_{P/IC}+\bomega\times(\bomega\times{\bf r}_{P/IC}).
\end{align}

**Remarks:** the IC generally migrates through space; for a translating rigid body the IC is at infinity; the IC can only be used for velocity analysis since its acceleration is nonzero.

**Procedure:** use geometry to find the IC; find $|\omega|$ via ${\bf v}_P = \bomega\times{\bf r}_{P/IC}$; obtain ${\bf v}_P$ for any point.

*Examples: [Crank slider](https://www.youtube.com/watch?v=RDixpdSGERQ), rod falling on a corner.*

## Doing Kinematics Geometrically

In some problems, geometric laws that hold for all time can be differentiated:
- The law of sines holds at all times.
- The law of cosines holds at all times.
- Angle sum relations for a triangle (e.g. $\beta+\alpha+\theta=\pi$) can be differentiated.

## Kinematics of Rolling and Sliding

::: {.callout-important title="Note!"}
Terminology: "rolling" = roll without slip; "sliding" = roll with slip.
:::

Consider a rigid body $\mathcal{B}$ in contact with a fixed surface $\mathcal{S}$. Let $P$ be the material point of $\mathcal{B}$ in contact at time $t$, with unit normal ${\bf n}$ to $\mathcal{S}$.

Since $P$ belongs to $\mathcal{B}$:
\begin{align}
    {\bf v}_P-{\bf v}_C &= \bomega\times({\bf r}_P-{\bf r}_C),\\
    {\bf a}_P-{\bf a}_C &= \balpha\times({\bf r}_P-{\bf r}_C)+\bomega\times\lp\bomega\times({\bf r}_P-{\bf r}_C)\rp.
\end{align}

- **Contact (touching):** ${\bf v}_P\cdot{\bf n} = 0$.
- **Rolling (no slip):** ${\bf v}_P = {\bf 0}$, so ${\bf v}_C = -\bomega\times({\bf r}_P-{\bf r}_C)$. Note: ${\bf a}_P \neq {\bf 0}$ in general.

### Example: Rolling Circular Disk

::: {.callout-tip title="Think!"}
**Question:** What is the IC of a rolling disk?

::: {.callout-note title="Answer" collapse="true"}
The contact point with the surface.
:::
:::

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw (-1,0) -- (7,0);
    \draw (3,3) circle (3);
    \draw[->] (3,3) -- (5,3) node[right] {${\bf E}_x$};
    \draw[->] (3,3) -- (3,5) node[above] {${\bf E}_y$};
    \draw[->] (3,3) -- (4.8,4) node[right] {${\bf e}_x$};
    \draw[->] (3,3) -- (2,4.7) node[above] {${\bf e}_y$};
    \draw[->] (-4,2) -- (-2,2) node[right] {${\bf E}_x$};
    \draw[->] (-4,2) -- (-4,4) node[above] {${\bf E}_y$};
    \node at (-4.2,1.8) {$O$};
    \draw[->] (-4,2) -- (3,0) node[midway,below] {${\bf r}_P$};
    \draw[->] (-4,2) -- (3,3) node[midway,above] {${\bf r}_C$};
    \node at (3,2.7) {$C$};
    \node at (3,-0.3) {$P$};
\end{tikzpicture}
```

A disk of radius $R$ rolling on a plane with $\bomega = \omega{\bf E}_z$. Since ${\bf v}_P = {\bf 0}$:
\begin{align}
    {\bf v}_C = \bomega\times R{\bf E}_y = -\omega R{\bf E}_x \implies \dot{x} = -\omega R, \quad \ddot{x} = -\alpha R.
\end{align}
The acceleration of the contact point:
\begin{align}
    {\bf a}_P = -\omega^2 R{\bf E}_y
\end{align}
(purely vertical — toward the center).

## Center of Mass

The center of mass $C$ of $\mathcal{B}$ has position vector
\begin{align}
    {\bf r}_C = \frac{\int_{\mathcal{B}}{\bf r}\,dm}{\int_{\mathcal{B}}dm}, \qquad dm = \rho(x,t)\,dv.
\end{align}
For rigid bodies, $C$ behaves as a material point:
\begin{align}
    {\bf v}_C-{\bf v}_A = \bomega\times({\bf r}_C-{\bf r}_A), \qquad
    {\bf a}_C-{\bf a}_A = \balpha\times({\bf r}_C-{\bf r}_A)+\bomega\times\lp\bomega\times({\bf r}_C-{\bf r}_A)\rp.
\end{align}

## Linear Momentum

\begin{align}
    {\bf G} = \int_{\mathcal{B}}{\bf v}\,dm = m{\bf v}_C.
\end{align}

## Angular Momenta

The angular momentum about a point $P$ is
\begin{align}
    {\bf H}^P = \int_{\mathcal{B}}({\bf r}-{\bf r}_P)\times{\bf v}\,dm,
\end{align}
and the relationship between angular momenta about $O$ and $C$ is
\begin{align}
    {\bf H}^O = {\bf H}^C+{\bf r}_C\times{\bf G}.
\end{align}

### Inertia Tensor

With $\bpi = {\bf r}-{\bf r}_C$ and using the BAC-CAB identity:
\begin{align}
    {\bf H}^C = \int_{\mathcal{B}}\lp(\bpi\cdot\bpi)\bomega-(\bpi\cdot\bomega)\bpi\rp dm = {\bf I}^C\bomega.
\end{align}
The inertia matrix on the corotational basis:
\begin{align}
    [{\bf I}^C] =
    \begin{bmatrix}
        I^C_{xx} & I^C_{xy} & I^C_{xz}\\
        I^C_{xy} & I^C_{yy} & I^C_{yz}\\
        I^C_{xz} & I^C_{yz} & I^C_{zz}
    \end{bmatrix},
\end{align}
with diagonal entries (**moments of inertia**) and off-diagonal entries (**products of inertia**):
\begin{align}
    I^C_{xx} &= \int_{\mathcal{B}}(y^2+z^2)\,dm, \qquad
    I^C_{yy} = \int_{\mathcal{B}}(x^2+z^2)\,dm, \qquad
    I^C_{zz} = \int_{\mathcal{B}}(x^2+y^2)\,dm,\\
    I^C_{xy} &= -\int_{\mathcal{B}}xy\,dm, \qquad
    I^C_{xz} = -\int_{\mathcal{B}}xz\,dm, \qquad
    I^C_{yz} = -\int_{\mathcal{B}}yz\,dm.
\end{align}
When $\{{\bf e}_x,{\bf e}_y,{\bf e}_z\}$ is an eigenbasis of $[{\bf I}^C]$, the products of inertia vanish.

::: {.callout-tip title="Think!"}
**Question:** Is my moment of inertia about an axis through my body greater when my arms are closed or open?

::: {.callout-note title="Answer" collapse="true"}
Greater when arms are open — mass is farther from the axis. See also: [Professor Walter Lewin rolling cylinders](https://www.youtube.com/watch?v=M_YCWDXCwZM).
:::
:::

### Example: Solid Cylinder

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}[scale=2]
    \draw[thick] (0,1) ellipse (0.5 and 0.2);
    \draw[thick] (-0.5,1) -- (-0.5,-1);
    \draw[thick] (0.5,1) -- (0.5,-1);
    \draw[thick, dashed] (0,-1) ellipse (0.5 and 0.2);
    \draw[->, thick, blue] (0,0) -- (0,0.5) node[anchor=south] {${\bf E}_z$};
    \draw[->, thick, blue] (0,0) -- (-0.4,-0.2) node[anchor=north] {${\bf E}_x$};
    \draw[->, thick, blue] (0,0) -- (0.4,-0.2) node[anchor=north] {${\bf E}_y$};
    \fill (0,0) circle (0.04);
    \node at (0,-0.15) {$C$};
    \draw[<->] (-1,-1) -- (-1,1) node[midway,left] {$\ell$};
    \draw[->] (0,-0.5) -- (0.5,-0.5) node[midway,below] {$R$};
\end{tikzpicture}
```

With $dm = \rho\,r\,dr\,d\theta\,dz$:
\begin{align}
    I^C_{zz} = \rho\int_{-\ell/2}^{\ell/2}\int_0^{2\pi}\int_0^R r^3\,dr\,d\theta\,dz = \frac{1}{2}\rho\pi\ell R^4 = \frac{mR^2}{2}.
\end{align}

### Example: Cylindrical Hoop

Every $dm$ has $r = R$, so:
\begin{align}
    I^C_{zz} = R^2\int_{\mathcal{B}}dm = mR^2.
\end{align}

## Parallel Axis Theorem

Given inertia components about $C$, the inertia about a material point $A$ with ${\bf r}_A - {\bf r}_C = A_x{\bf e}_x+A_y{\bf e}_y+A_z{\bf e}_z$ (axes parallel at $A$ and $C$):
\begin{align}
    I^A_{xx} &= I^C_{xx}+m(A_y^2+A_z^2),\\
    I^A_{xy} &= I^C_{xy}-mA_xA_y.
\end{align}
In tensor form:
\begin{align}
    {\bf I}^A = {\bf I}^C+m\lnorm{\bf r}_A-{\bf r}_C\rnorm^2{\bf I}-m({\bf r}_A-{\bf r}_C)\otimes({\bf r}_A-{\bf r}_C).
\end{align}

::: {.callout-important title="Note!"}
The parallel axis theorem applies only between the center of mass $C$ and another material point — not between any two arbitrary material points. The moment of inertia always increases when moving away from $C$.
:::

## Radius of Gyration

The radius of gyration $k_z$ about the $z$-axis satisfies $mk_z^2 = I_{zz}$.

## Angular Momentum About a Moving Point $P$

For a material point $P$ on the rigid body:
\begin{align}
    {\bf H}^P &= {\bf I}^P\bomega+({\bf r}_C-{\bf r}_P)\times m{\bf v}_P,
\end{align}
with special cases ${\bf H}^C = {\bf I}^C\bomega$ and ${\bf H}^O = {\bf I}^O\bomega$ (fixed $O$). In general:
\begin{align}
    \dot{\bf H}^P = {\bf I}^P\overset{\circ}{\bomega}+\bomega\times({\bf I}^P\bomega)+\frac{d}{dt}\leb({\bf r}_C-{\bf r}_P)\times m{\bf v}_P\reb.
\end{align}


## Summary

**Corotational basis** $\{\mathbf{e}_x,\mathbf{e}_y,\mathbf{E}_z\}$ fixed to the rigid body:
\begin{align}
    \dot{\mathbf{e}}_x = \boldsymbol{\omega}\times\mathbf{e}_x, \qquad
    \dot{\mathbf{e}}_y = \boldsymbol{\omega}\times\mathbf{e}_y, \qquad
    \boldsymbol{\omega}=\dot\theta\,\mathbf{E}_z.
\end{align}

**Velocity/acceleration of two points** $A$, $B$ on a rigid body:
\begin{align}
    \mathbf{v}_B - \mathbf{v}_A &= \boldsymbol{\omega}\times(\mathbf{r}_B-\mathbf{r}_A), \\
    \mathbf{a}_B - \mathbf{a}_A &= \boldsymbol{\alpha}\times(\mathbf{r}_B-\mathbf{r}_A)+\boldsymbol{\omega}\times(\mathbf{v}_B-\mathbf{v}_A).
\end{align}

**Instantaneous center (IC)**: the point on a body (or its extension) with instantaneous zero velocity. Lies at the intersection of perpendiculars to the velocity vectors of two points.

**Particle in a non-inertial frame** (point $A$ fixed to body, $B$ moving relative to body):
\begin{align}
    \mathbf{v}_B - \mathbf{v}_A &= \boldsymbol{\omega}\times(\mathbf{r}_B-\mathbf{r}_A)+\mathbf{v}_{\mathrm{rel}}, \\
    \mathbf{a}_B - \mathbf{a}_A &= \boldsymbol{\alpha}\times(\mathbf{r}_B-\mathbf{r}_A)+\boldsymbol{\omega}\times(\boldsymbol{\omega}\times(\mathbf{r}_B-\mathbf{r}_A))+2\boldsymbol{\omega}\times\mathbf{v}_{\mathrm{rel}}+\mathbf{a}_{\mathrm{rel}}.
\end{align}
The Coriolis acceleration is $2\boldsymbol{\omega}\times\mathbf{v}_{\mathrm{rel}}$.

## Exercises

### Set 15 – Rigid Body Kinematics

**1.** [MKB 05-010] *(ans. $\mathbf{v}_B=-11\mathbf{E}_x$ m/s, $\mathbf{a}_B=22\mathbf{E}_x-220\mathbf{E}_y$ m/s$^2$)*

**2.** [MKB 05-013] *(ans. $\mathbf{v}_A=1.121\mathbf{E}_x+0.838\mathbf{E}_y$ m/s, $\mathbf{a}_B=-4.48\mathbf{E}_x+0.1465\mathbf{E}_y$ m/s$^2$)*

**3.** [05-018] *(ans. $\mathbf{v}_A=-0.374\mathbf{E}_x+0.1905\mathbf{E}_y$ m/s, $\mathbf{a}_A=-0.757\mathbf{E}_x-0.605\mathbf{E}_y$ m/s$^2$)*

**4.** [05-049] *(ans. (a) $N=91.7$ rev/min CCW; (b) $N=45.8$ rev/min CCW; (c) $N=45.8$ rev/min CW)*

**5.** [05-065] *(ans. $\omega=0.722$ rad/s)*

**6.** [05-069] *(ans. $\omega_{AB}=1.725$ rad/s CCW, $\omega_{BC}=4$ rad/s CCW)*

### Set 16 – IC and Particle in Non-inertial Frame

**1.** [MKB 05-102] *(ans. $\alpha=0.286$ rad/s$^2$, $a_A=0.653$ m/s$^2$ down)*

**2.** [MKB 05-125] *(ans. $\mathbf{v}_A=0.1\mathbf{E}_x+0.25\mathbf{E}_y$ m/s, $\beta=68.2^\circ$)*

**3.** [05-143] *(ans. $v_{\mathrm{rel}}=3.93$ m/s, $a_{\mathrm{rel}}=15.22$ m/s$^2$, $\omega_{BC}=1.046$ rad/s CW, $\alpha_{BC}=117.7$ rad/s$^2$ CW)*

### Set 17 – Rolling and Sliding

**1.** [MKB 05-055] *(ans. $v_O=6.93$ m/s, $\omega=21.3$ rad/s CW)*

**2.** [MKB 05-087] *(ans. $v_O=120$ mm/s, $v_P=216$ mm/s)*

**3.** [05-091] *(ans. $v=15.71$ ft/s right, $v_s=6.89$ ft/s left)*

**4.** [05-093] *(ans. $v_A=9.19$ ft/sec left)*

**5.** [05-100] *(ans. $\mathbf{v}_A=5.12\mathbf{E}_x+2.12\mathbf{E}_y$ m/s, $\mathbf{a}_B=-16.25\mathbf{E}_x+2.5\mathbf{E}_y$ m/s$^2$)*
