---
title: "Kinetics of Rigid Bodies"
---

## Balance Laws for a Rigid Body

The balance laws for a rigid body are **Euler's first law** (Balance of Linear Momentum, BoLM) and **Euler's second law** (Balance of Angular Momentum, BoAM):
\begin{align}
    {\bf F} &= m{\bf a}_C,\\
    {\bf M}^O &= \dot{\bf H}^O \quad\text{about a fixed point }O,\\
    {\bf M}^C &= \dot{\bf H}^C \quad\text{about the center of mass }C,\\
    {\bf M}^P &= \dot{\bf H}^P+({\bf v}_P-{\bf v}_C)\times{\bf G} \quad\text{about any material point }P.
\end{align}
The BoAM is an **independent postulate**, not derivable from the BoLM.

## Resultant Forces and Moments

The resultant force ${\bf F}$ is the sum of all external forces. The resultant moment about a point includes the moment due to forces plus any applied pure moments ${\bf M}_e$ not due to forces.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}[scale=1]
    \draw[thick] plot [smooth cycle, tension=1] coordinates {
        (0,0) (2,3) (5,2.5) (7,1) (4,-2) (1,-3)};
    \fill[blue] (1,0.5) circle (1.5pt) node[anchor=north west] {$X_1$};
    \fill[blue] (6,0) circle (1.5pt) node[anchor=south] {$X_2$};
    \fill[blue] (2,-2) circle (1.5pt) node[anchor=north east] {$X_3$};
    \fill[black] (-2,-2) circle (1.5pt) node[anchor=north east] {$O$};
    \draw[thick,->] (-2,-2) -- (1,0.5) node[midway,above] {${\bf r}_1$};
    \draw[thick,->] (-2,-2) -- (6,0) node[midway,above] {${\bf r}_2$};
    \draw[thick,->] (-2,-2) -- (2,-2) node[midway,below] {${\bf r}_3$};
    \node[anchor=south east] at (1,2.2) {$\mathcal{B}$};
    \draw[->, thick, blue] (1,0.5) -- ++(0.8,1.2) node[anchor=west] {$\mathbf{F}_1$};
    \draw[->, thick, blue] (6,0) -- ++(-0.8,0.4) node[anchor=south] {$\mathbf{F}_2$};
    \draw[->, thick, blue] (2,-2) -- ++(0.8,-1.2) node[anchor=north] {$\mathbf{F}_3$};
    \draw[->, thick, blue] (4,2) arc[start angle=0, end angle=260, radius=0.4]
        node[right] {$\mathbf{M}_e$};
    \fill[red] (3,0) circle (1.5pt) node[anchor = south east] {$C$};
\end{tikzpicture}
```

For $K$ forces ${\bf F}_i$ acting at positions ${\bf r}_i$ plus a pure moment ${\bf M}_e$:
\begin{align}
    {\bf F} &= \sum_{i=1}^K{\bf F}_i,\\
    {\bf M}^O &= {\bf M}_e+\sum_{i=1}^K{\bf r}_i\times{\bf F}_i,\\
    {\bf M}^C &= {\bf M}_e+\sum_{i=1}^K({\bf r}_i-{\bf r}_C)\times{\bf F}_i.
\end{align}

Examples of pure moments: reaction moments ${\bf M}_R$ at joints; torsional spring moments $-K_T(\theta-\theta_0){\bf E}_z$.

### Does Weight Give a Moment About the Center of Mass?

The total weight is ${\bf W} = \int_{\mathcal{B}}{\bf g}\,dm = m{\bf g}$, and its moment about $C$ is:
\begin{align}
    {\bf M}^C = \int_{\mathcal{B}}\bpi\times{\bf g}\,dm = \lp\int_{\mathcal{B}}\bpi\,dm\rp\times{\bf g} = {\bf 0},
\end{align}
since $\int_{\mathcal{B}}\bpi\,dm = {\bf 0}$ by definition of the center of mass.

## Equivalence of the BoAM Forms

### BoAM About Any Point $P$

Starting from ${\bf M}^C = \dot{\bf H}^C$ and replacing ${\bf M}^P = {\bf M}^C+({\bf r}_C-{\bf r}_P)\times{\bf F}$:
\begin{align}
    {\bf M}^P = \dot{\bf H}^C+({\bf r}_C-{\bf r}_P)\times{\bf F}.
\end{align}
Substituting $\dot{\bf H}^C = \dot{\bf H}^P+({\bf v}_P-{\bf v}_C)\times{\bf G}+({\bf r}_P-{\bf r}_C)\times m{\bf a}_C$ and simplifying:
\begin{align}
    {\bf M}^P = \dot{\bf H}^P+({\bf v}_P-{\bf v}_C)\times{\bf G},
\end{align}
recovering the BoAM about a general material point $P$.

{{< video https://www.youtube.com/shorts/VmeM0BNnGR0 title="Ice skater conservation of angular momentum" >}}

{{< video https://www.youtube.com/watch?v=64t-dVtDwkQ title="Angular momentum conservation on Hoberman Sphere - Dr. Boyd F. Edwards" >}}

### BoAM About a Fixed Point $O$

If $P$ is a fixed point $O$ (so ${\bf v}_O = {\bf 0}$), the above simplifies to ${\bf M}^O = \dot{\bf H}^O$.

## Calculating the Derivative of Angular Momentum

Recall ${\bf H}^C = {\bf I}^C\bomega$ and $\dot{\bf H}^C = \overset{\circ}{\bf H}^C+\bomega\times{\bf H}^C$, where $\overset{\circ}{\bf H}^C$ is the corotational rate (time derivative with ${\bf e}_x,{\bf e}_y,{\bf e}_z$ fixed).

For $\bomega = \omega{\bf E}_z$ (the case exclusively considered in this class):
\begin{align}
    {\bf H}^C &= I^C_{xz}\omega_z{\bf e}_x+I^C_{yz}\omega_z{\bf e}_y+I^C_{zz}\omega_z{\bf e}_z,\\
    \dot{\bf H}^C &= (I^C_{xz}\dot{\omega}-I^C_{yz}\omega^2){\bf e}_x+(I^C_{yz}\dot{\omega}+I^C_{xz}\omega^2){\bf e}_y+I^C_{zz}\dot{\omega}{\bf E}_z.
\end{align}
Analogously for other reference points.

::: {.callout-important title="Note!"}
For fixed axis rotation, sum moments about a fixed point on the axis. For general plane motion, sum about $C$ or a general material point $P$ using
\begin{align}
    {\bf M}^P = (I^C_{xz}\dot{\omega}-I^C_{yz}\omega^2){\bf e}_x+(I^C_{yz}\dot{\omega}+I^C_{xz}\omega^2){\bf e}_y+I^C_{zz}\dot{\omega}{\bf E}_z+({\bf r}_C-{\bf r}_P)\times m{\bf a}_C.
\end{align}
In most problems, axes are chosen so that $I_{xz}=I_{yz}=0$.
:::

## Conservative Moments

### Torsional Spring

A torsional spring with stiffness $K$ creates a couple ${\bf M}_S = -K\theta{\bf E}_z$. Its power is:
\begin{align}
    {\bf M}_S\cdot\bomega = -K\theta\dot{\theta} = -\frac{d}{dt}\lp\frac{K\theta^2}{2}\rp.
\end{align}
Hence $U = \frac{1}{2}K\theta^2$ for the torsional spring.

### Constant Couple

A constant couple ${\bf M}_e = M_e{\bf E}_z$ has power ${\bf M}_e\cdot\bomega = M_e\dot{\theta} = -\frac{d}{dt}(-M_e\theta)$. Hence $U = -M_e\theta$.

::: {.callout-important title="Note!"}
These results apply only for $\bomega = \omega{\bf E}_z$. Constant couples are not conservative in general.
:::

## Impact Problems and Center of Percussion

For rigid bodies, impact ideas are the same as in systems of particles.

*Example:* Consider a projectile fired horizontally at a pendulum (in the horizontal plane, no gravity). Find position $x$ along the pendulum such that the reaction force at the pin is nearly zero during impact.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick] (0,0) -- (0,-4);
    \fill (0,0) circle (0.1);
    \fill (-2,-3) circle (0.25);
    \draw[thick,->] (-2,-3) -- (-1,-3) node[above] {${\bf v}_0$};
    \draw[->] (1,0) -- (1,-3) node[midway, right] {$x$};
\end{tikzpicture}
```

If $x$ is large, the horizontal reaction at $O$ points left; if $x$ is small, it points right. The optimal $x$ that minimizes the reaction is called the **center of percussion**.

## Summary

For $K$ forces and a moment ${\bf M}_e$ acting on a rigid body, the balance laws are:
\begin{align}
    {\bf F} = m\frac{d{\bf v}_C}{dt}, \qquad {\bf M}^O = \dot{\bf H}^O \quad (\text{or equivalently } {\bf M}^C = \dot{\bf H}^C).
\end{align}

For fixed-axis rotation with $\bomega = \omega{\bf E}_z$ and $I_{xz}=I_{yz}=0$:
\begin{align}
    {\bf F} = m\dot{\bf v}_C, \qquad {\bf M} = I_{zz}\dot{\omega}{\bf E}_z.
\end{align}

Four classes of applications: (1) purely translational motion ($\bomega=\balpha={\bf 0}$), (2) rigid body with a fixed point, (3) rolling and sliding rigid bodies, (4) imbalanced rotors.


## Summary

**BoLM**: $\mathbf{F}=m\mathbf{a}_C$.

**BoAM** (three equivalent forms):
\begin{align}
    \mathbf{M}^O = \dot{\mathbf{H}}^O \quad &(\text{fixed point } O), \\
    \mathbf{M}^C = \dot{\mathbf{H}}^C \quad &(\text{center of mass } C), \\
    \mathbf{M}^P = \dot{\mathbf{H}}^C + (\mathbf{r}_C-\mathbf{r}_P)\times m\mathbf{a}_C \quad &(\text{any material point } P).
\end{align}

**Angular momentum and inertia tensor**:
\begin{align}
    H^C_z = I^C_{zz}\,\omega, \qquad
    I^C_{zz} = \int_{\mathcal{B}}(x^2+y^2)\,dm.
\end{align}

**Parallel axis theorem**: $I^A_{zz} = I^C_{zz} + m d^2$ where $d$ is the distance from $C$ to $A$.


## Exercises

### Set 18 – Moments of Inertia

**1.** [MKB B-004] Integration required. *(ans. $I_{xx}=\frac{3}{10}mr^2$, $I_{yy}=\frac{2}{5}m(r^4+h^2)/(r^2)$)*

**2.** [MKB B-029] Treat the hollow cylinder as a full cylinder of radius $r_2$ minus a cylinder of radius $r_1$. *(ans. $I_{xx}=\frac{1}{2}m(r_1^2+r_2^2)$)*

**3.** [B-032] *(ans. $L=r\sqrt{2\sqrt{3}}$)*

**4.** [B-034] *(ans. $I_{yy}=\rho L^3\left(\frac{43}{192}+\frac{83\pi}{128}\right)$)*

### Set 19 – Rigid Body Translation

**1.** [MKB 06-006] *(ans. $a=5.66$ m/s$^2$)*

**2.** [06-012] *(ans. $a=16.43$ ft/sec$^2$)*

**3.** [06-013] *(ans. $N_A=6.85$ kN up, $N_B=9.34$ kN up)*

**4.** [06-022] *(ans. $N=257$ kN up)*

### Set 20 – Fixed Point Rotation

**1.** [06-051] *(See problem set for figure.)*

{{< video https://youtu.be/Q7PV2pxBuwc title="Solution – Problem 06/051" width="100%" >}}


### Set 21 – General Plane Motion

**1.** [MKB 06-061] *(ans. $\alpha=48.8$ rad/s$^2$ CW, $\bar a_x=0$, $\bar a_y=5$ m/s$^2$)*

**2.** [MKB 06-062] *(ans. A: $\alpha_A=\frac{g}{r}\sin\theta$, $\mu_s=0$; B: $\alpha_B=\frac{g}{2r}\sin\theta$, $\mu_s=\frac{1}{2}\tan\theta$)*

**3.** [06-063] *(ans. $\bar a=13.80$ ft/sec$^2$ down incline, $F=1.714$ lb up incline)*

**4.** [06-070] *(ans. $a=\frac{8(m+M)g}{3\pi(m+3M)}$ left, $\alpha=\frac{8mg}{3\pi r(m+3M)}$ CW)*

{{< video https://youtu.be/0_wB2JnICLc title="Solution – Problem 06/070" width="100%" >}}


**5.** [06-076] *(ans. $s=\frac{3d}{2}$)*

**6.** [06-077] *(ans. $N_B=36.4$ N up)*

### Set 22 – Impulse and Momentum for Rigid Bodies

**1.** [MKB 06-136] *(ans. $\omega=1.811$ rad/s CCW)*

**2.** [MKB 06-142] *(ans. $\mathbf{v}=\frac{Mu_M\mathbf{E}_x+m\mathbf{v}_m}{M+m}$, $\omega=\frac{12vm}{L(4M+7m)}$ CCW)*

**3.** [06-145] *(ans. $\omega=\frac{3mv_1}{(M+m)L}$ CW)*

**4.** [06-146] *(ans. $N=2.04$ rev/s)*

**5.** [06-148] *(ans. $\omega=1.593$ rad/s CCW, $n=91.7\%$)*

**6.** [06-155] *(ans. $t=\frac{2v_0}{g(7\mu_k\cos\theta-2\sin\theta)}$)*

{{< video https://youtu.be/0N8tDgnrXNA title="Solution – Problem 06/155" width="100%" >}}

