---
title: "The Work-Energy Theorem for a Rigid Body"
---

The **Koenig decomposition** for the kinetic energy of a rigid body is:
\begin{align}
    T = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}{\bf H}^C\cdot\bomega.
\end{align}
In this class, $\bomega = \omega{\bf E}_z$, so $\frac{1}{2}{\bf H}^C\cdot\bomega = \frac{1}{2}I_{zz}\omega^2$.

For fixed point rotation about $O$, the kinetic energy simplifies to:
\begin{align}
    T = \frac{1}{2}{\bf H}^O\cdot\bomega = \frac{1}{2}I^O_{zz}\omega^2.
\end{align}

The **work-energy theorem** for a rigid body is:
\begin{align}
    \frac{dT}{dt} = {\bf F}\cdot{\bf v}_C+{\bf M}^C\cdot\bomega = \sum_{i=1}^N{\bf F}_i\cdot{\bf v}_i+{\bf M}_e\cdot\bomega.
\end{align}
Integrating from $t_A$ to $t_B$:
\begin{align}
    T_B-T_A = W_{{\bf F},AB}+W_{{\bf M},AB} = \sum_{i=1}^N W_{{\bf F}_i,AB}+W_{{\bf M}_e,AB},
\end{align}
where
\begin{align}
    W_{{\bf F},AB} &= \int_{t_A}^{t_B}{\bf F}\cdot{\bf v}_C\,dt, &
    W_{{\bf M},AB} &= \int_{t_A}^{t_B}{\bf M}^C\cdot\bomega\,dt,\\
    W_{{\bf F}_i,AB} &= \int_{t_A}^{t_B}{\bf F}_i\cdot{\bf v}_i\,dt, &
    W_{{\bf M}_e,AB} &= \int_{t_A}^{t_B}{\bf M}_e\cdot\bomega\,dt.
\end{align}
Notice: the work of a force uses the velocity of its **point of application**.

## Koenig's Decomposition

By definition, the kinetic energy of a rigid body is:
\begin{align}
    T = \frac{1}{2}\int_{\mathcal{B}}{\bf v}\cdot{\bf v}\,\rho\,dv.
\end{align}
Using ${\bf v} = {\bf v}_C+\bomega\times\bpi$ with $\bpi = {\bf r}-{\bf r}_C$:
\begin{align}
    T = \frac{1}{2}\int_{\mathcal{B}}\lp{\bf v}_C\cdot{\bf v}_C+2{\bf v}_C\cdot(\bomega\times\bpi)+(\bomega\times\bpi)\cdot(\bomega\times\bpi)\rp dm.
\end{align}
The first term integrates to $\frac{1}{2}m{\bf v}_C\cdot{\bf v}_C$, the second vanishes (since $\int\bpi\,dm = {\bf 0}$), and using the BAC-CAB identity for the third:
\begin{align}
    T = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}{\bf H}^C\cdot\bomega.
\end{align}

### Fixed Point Rotation

For fixed point rotation about $O$ (${\bf v}_O = {\bf 0}$), using the cyclic property of the scalar triple product:
\begin{align}
    T &= \frac{1}{2}{\bf G}\cdot(\bomega\times{\bf r}_{C/O})+\frac{1}{2}{\bf H}^C\cdot\bomega
    = \frac{1}{2}({\bf r}_{C/O}\times{\bf G}+{\bf H}^C)\cdot\bomega
    = \frac{1}{2}{\bf H}^O\cdot\bomega.
\end{align}

## Derivation of the Work-Energy Theorem

From $T = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}{\bf H}^C\cdot\bomega$, taking the time derivative and using the fact that $\dot{\bf H}^C\cdot\bomega = {\bf H}^C\cdot\dot{\bomega}$ (verified by direct calculation):
\begin{align}
    \dot{T} = m\dot{\bf v}_C\cdot{\bf v}_C+\dot{\bf H}^C\cdot\bomega.
\end{align}
Invoking the BoLM and BoAM:
\begin{align}
    \dot{T} = {\bf F}\cdot{\bf v}_C+{\bf M}^C\cdot\bomega.
\end{align}

## Alternative Form

Using ${\bf F} = \sum_i{\bf F}_i$ and ${\bf M}^C = {\bf M}_e+\sum_i({\bf r}_i-{\bf r}_C)\times{\bf F}_i$, and noting that ${\bf v}_i = {\bf v}_C+\bomega\times({\bf r}_i-{\bf r}_C)$:
\begin{align}
    \dot{T} = \sum_{i=1}^K{\bf F}_i\cdot{\bf v}_i+{\bf M}_e\cdot\bomega.
\end{align}
This form is often easier to use in applications: each force is dotted with the velocity of its point of application.

## Examples

### Bar Pendulum Fixed at Its End

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick] (0,0) -- (3,-4);
    \fill[black] (0,0) circle (0.1);
    \draw[thick,->] (0,0) -- (0,-1) node[below] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (1,0) node[right] {${\bf E}_y$};
    \draw[thick,->,blue] (0,0) -- (0.6,-0.8) node[below] {${\bf e}_x$};
    \draw[thick,->,blue] (0,0) -- (0.8,0.6) node[above] {${\bf e}_y$};
\end{tikzpicture}
```

A bar of length $\ell$ and mass $m$ pinned at its end. The energy is conserved:
\begin{align}
    U &= -\frac{mg\ell}{2}\cos\theta,\\
    E &= \frac{m\ell^2}{6}\dot{\theta}^2-\frac{mg\ell}{2}\cos\theta = \text{const}.
\end{align}

### Bar Pendulum With End Mass

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick] (0,0) -- (3,-4);
    \fill[black] (0,0) circle (0.1);
    \draw[thick,->] (0,0) -- (0,-1) node[below] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (1,0) node[right] {${\bf E}_y$};
    \draw[thick,->,blue] (0,0) -- (0.6,-0.8) node[below] {${\bf e}_x$};
    \draw[thick,->,blue] (0,0) -- (0.8,0.6) node[above] {${\bf e}_y$};
    \fill[black] (3,-4) circle (0.4);
\end{tikzpicture}
```

Adding a dead mass $m_e$ at the end:
\begin{align}
    U = -\frac{mg\ell}{2}\cos\theta-m_e g\ell\cos\theta.
\end{align}

### Follower Force

If instead a follower force ${\bf P}$ is applied perpendicular to the bar at its end, this force is **nonconservative** (its direction actively changes with configuration). Energy is not conserved.

## Energy Conservation

As with particles and systems of particles, this theorem can be used to establish conservation of the total mechanical energy of a rigid body, $E = T + U$, when all work-doing forces and moments are conservative.


## Summary

**Koenig decomposition** of kinetic energy:
\begin{align}
    T = \tfrac{1}{2}m\mathbf{v}_C\cdot\mathbf{v}_C + \tfrac{1}{2}\mathbf{H}^C\cdot\boldsymbol{\omega}.
\end{align}
If a point $O$ has zero velocity: $T = \tfrac{1}{2}\mathbf{H}^O\cdot\boldsymbol{\omega}$.

**Work–energy theorem for a rigid body**:
\begin{align}
    \frac{dT}{dt} = \mathbf{F}\cdot\mathbf{v}_C + \mathbf{M}\cdot\boldsymbol{\omega} = \sum_i \mathbf{F}_i\cdot\mathbf{v}_i + M_e\omega.
\end{align}

## Lecture Videos

{{< video https://youtu.be/acmYFwl8rRo title="Introducing the work–energy theorem for a rigid body" width="70%" >}}

{{< video https://youtu.be/3HRYCW7wiRs title="Deriving the kinetic energy of a rigid body" width="70%" >}}

{{< video https://youtube.com/shorts/3bu7kY8JLpM title="Kinetic energy – rigid body rotating about a fixed point" width="70%" >}}

{{< video https://youtu.be/vCROTYHJRow title="Deriving the work–energy theorem for a rigid body" width="70%" >}}


## Exercises

*The following problems are from Set 23 – Work–Energy Theorem for Rigid Bodies.*

**1.** [MKB 06-096] *(ans. $v_A=\sqrt{2gx\sin\theta}$, $v_B=\sqrt{gx\sin\theta}$)*

**2.** [MKB 06-099] *(ans. $\omega_{\max}=0.861\sqrt{g/b}$)*

**3.** [06-108] *(ans. (a) $k=93.3$ N/m; (b) $\omega=1.484$ rad/s CW)*

**4.** [06-112] *(ans. $x=0.211l$, $\omega_{\max}=1.861\sqrt{g/l}$ CW)*

**5.** [06-118] *(ans. $v_A=\sqrt{3}\sqrt{\frac{M\theta}{m}-gb(1-\cos\theta)}$ right)*

{{< video https://youtu.be/-jdslhyHAIY title="Solution – Problem 06/118" width="70%" >}}

