---
title: "System of Particles"
---

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 5
\begin{tikzpicture}
    \draw[thick,->] (0,0) -- (0,1) node[above] {${\bf E}_z$};
    \draw[thick,->] (0,0) -- (-1.3,-1) node[left] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (1.3,-1) node[right] {${\bf E}_y$};
    \fill (0,0) circle (0.05);
    \node at (-0.3,0.1) {$O$};
    \fill[gray!40] (3,1) circle (0.3);
    \node at (2.8,1.5) {$m_i$};
    \draw[thick,->] (3,1) -- (5,0) node[right] {${\bf v}_i$};
    \draw[thick,->,blue] (0,0) -- (3,1) node[midway,above] {${\bf r}_i$};
    \fill[gray!40] (-3,2) circle (0.3);
    \node at (-3,2.5) {$m_1$};
    \draw[thick,->] (-3,2) -- (-4,1) node[left] {${\bf v}_1$};
    \draw[thick,->,blue] (0,0) -- (-3,2) node[midway,above] {${\bf r}_1$};
    \fill[gray!40] (1,-3) circle (0.3);
    \node at (1,-3.5) {$m_n$};
    \draw[thick,->] (1,-3) -- (2,-2) node[right] {${\bf v}_n$};
    \draw[thick,->,blue] (0,0) -- (1,-3) node[midway,left] {${\bf r}_n$};
\end{tikzpicture}
```

Consider a system of $n$ particles. A typical particle has mass $m_i$, position vector ${\bf r}_i$ relative to $O$, velocity ${\bf v}_i = \dot{\bf r}_i,$ acceleration ${\bf a}_i = \dot{\bf v}_i$, linear momentum ${\bf G}_i = m_i{\bf v}_i,$ angular momentum relative to some point $P$ ${\bf H}^P_i = ({\bf r}_i-{\bf r}_P)\times{\bf G}_i,$ and kinetic energy $T_i = \frac{1}{2}m_i{\bf v}_i\cdot{\bf v}_i$.

In this chapter, we define analogous system-level quantities and derive the balance laws for the system from the balance law for each individual particle.

## Center of Mass

::: {.callout-tip title="Think!"}
**Question:** What is the position vector of the center of mass of a system of particles?

::: {.callout-note title="Answer" collapse="true"}
The center of mass $C$ has position vector
\begin{align}
    {\bf r}_C = \frac{1}{m}\sum_{k=1}^n m_k{\bf r}_k,
\end{align}
where $m = \sum_{k=1}^n m_k$ is the total mass.
:::
:::

::: {.callout-warning title="Example"}
**Question:** Consider an unsymmetrical dumbbell: particle 1 of mass $m$ and particle 2 of mass $3m$ connected by a massless rigid rod, with
\begin{align*}
    {\bf r}_1 = {\bf E}_x+{\bf E}_y, \qquad {\bf r}_2 = 2{\bf E}_x.
\end{align*}
Calculate ${\bf r}_C$. Is $C$ located on the dumbbell?

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 2
\begin{tikzpicture}
    \fill[gray!50] (2,2) circle (0.2);
    \fill[gray!50] (4,0) circle (0.5);
    \fill (0,0) circle (0.05);
    \draw[thick] (2,2) -- (4,0);
    \draw[blue,thick,->] (0,0) -- (4,0) node[midway,below] {${\bf r}_2$};
    \draw[blue,thick,->] (0,0) -- (2,2) node[midway,above] {${\bf r}_1$};
    \draw[thick,->] (0,0) -- (1,0) node[below] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (0,1) node[above] {${\bf E}_y$};
    \node at (1.6,2) {$A$};
\end{tikzpicture}
```

::: {.callout-note title="Answer" collapse="true"}
\begin{align*}
    {\bf r}_C = \frac{m{\bf r}_1+3m{\bf r}_2}{m+3m} = \frac{{\bf r}_1+3{\bf r}_2}{4} = \frac{7}{4}{\bf E}_x+\frac{1}{4}{\bf E}_y.
\end{align*}
We can check mathematically that $C$ lies on the line connecting $m_1$ and $m_2$ by calculating the equation of that line and checking at $C$ belongs to it.

Alternatively, we can define a new basis, $\{{\bf u},{\bf v}\}$ and calculate the position of $C$ relative to $m_1$ on that basis. This should be along ${\bf u}$. 
:::
:::

Note the useful identity: relative to any reference point $A$,
\begin{align}
    {\bf r}_{C/A} = \frac{\sum_i m_i{\bf r}_{i/A}}{\sum_i m_i}.
\end{align}

::: {.callout-tip title="Think!"}
**Question:** Show that $\sum_{k=1}^n m_k({\bf r}_C-{\bf r}_k) = {\bf 0}$ and $\sum_{k=1}^n m_k({\bf v}_C-{\bf v}_k) = {\bf 0}$.

::: {.callout-note title="Answer" collapse="true"}
Both follow directly from the definition of the center of mass.
:::
:::

Differentiating the position of the center of mass:
\begin{align}
    {\bf v}_C = \frac{1}{m}\sum_{k=1}^n m_k{\bf v}_k = \frac{1}{m}\sum_{k=1}^n{\bf G}_k.
\end{align}
Notice that the velocity of the center of mass is the weighted sum of the velocities of the particles.

::: {.callout-tip title="Think!"}
**Question:** Show that $\sum_{k=1}^nm_k({\bf r}_C-{\bf r}_k) = {\bf 0}$ and $\sum_{k=1}^nm_k({\bf v}_C-{\bf v}_k) = {\bf 0}$.

::: {.callout-note title="Answer" collapse="true"}
...
:::
:::

## Linear Momentum

The linear momentum of the system is
\begin{align}
    {\bf G} = m\frac{d{\bf r}}{dt} = \sum_{k=1}^n m_k\frac{d{\bf r}_k}{dt} = \sum_{k=1}^n{\bf G}_k.
\end{align}
The linear momentum of the system equals that of its center of mass.

## Angular Momentum

The angular momentum ${\bf H}^P$ of the system of particles relative to a point $P$, whose position vector relative to $O$ is ${\bf r}_P$, is the sum of the individual angular momenta:
\begin{align}
    {\bf H}^P = \sum_{k=1}^n{\bf H}^{P_K} = \sum_{k=1}^n({\bf r}_k-{\bf r}_P)\times m_k{\bf v}_k = {\bf H}^C+({\bf r}_C-{\bf r}_P)\times{\bf G},
\end{align}
where
\begin{align}
    {\bf H}^C = \sum_{k=1}^n({\bf r}_k-{\bf r}_C)\times m_k{\bf v}_k = \sum_{k=1}^n({\bf r}_k-{\bf r}_C)\times m_k({\bf v}_k-{\bf v}_C).
\end{align}

::: {.callout-tip title="Think!"}
**Question:** Show that $\sum_{k=1}^n({\bf r}_k-{\bf r}_P)\times m_k{\bf v}_k = {\bf H}^C+({\bf r}_C-{\bf r}_P)\times{\bf G}$.

::: {.callout-note title="Answer" collapse="true"}
Introduce ${\bf r}_C-{\bf r}_C$ inside the sum and expand.
:::
:::

## Kinetic Energy

The kinetic energy of the system is:
\begin{align}
    T = \sum_{k=1}^n\frac{1}{2}m_k{\bf v}_k\cdot{\bf v}_k = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}\sum_{k=1}^n m_k({\bf v}_k-{\bf v}_C)\cdot({\bf v}_k-{\bf v}_C).
\end{align}
The kinetic energy splits into the kinetic energy of the center of mass plus the kinetic energy relative to the center of mass. In general, $T \neq \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C$.

::: {.callout-tip title="Think!"}
**Question:** Prove $\sum_{k=1}^n\frac{1}{2}m_k{\bf v}_k\cdot{\bf v}_k = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}\sum_{k=1}^n m_k({\bf v}_k-{\bf v}_C)\cdot({\bf v}_k-{\bf v}_C)$.

::: {.callout-note title="Answer" collapse="true"}
Add ${\bf v}_C-{\bf v}_C$ to each ${\bf v}_k$, expand, and use $\sum m_k({\bf v}_k-{\bf v}_C) = {\bf 0}$.
:::
:::

## Kinetics: Balance of Linear Momentum

Starting from ${\bf F}_i = m_i{\bf a}_i$ for each particle and summing:
\begin{align}
    {\bf F} = \sum_{k=1}^n{\bf F}_k = m{\bf a}_C.
\end{align}
The resultant external force equals the total mass times the acceleration of the center of mass.

::: {.callout-important title="Note!"}
The conditions for conservation of linear momentum of a system (completely or along a direction) are analogous to those for a single particle: ${\bf F}\cdot{\bf c} = 0$ for a constant ${\bf c}$.
:::

## Kinetics: Balance of Angular Momentum

Starting from ${\bf H}^P = \sum_{k=1}^n({\bf r}_k-{\bf r}_P)\times m_k{\bf v}_k$, the time derivative is:
\begin{align}
    \dot{\bf H}^P = \sum_{k=1}^n({\bf r}_k-{\bf r}_P)\times{\bf F}_k-{\bf v}_P\times{\bf G} = {\bf M}^P-{\bf v}_P\times{\bf G},
\end{align}
where the resultant moment of the system of forces relative to $P$ is ${\bf M}^P = \sum_{k=1}^n({\bf r}_k-{\bf r}_P)\times{\bf F}_k$.
The angular momentum theorem becomes
\begin{align}
    \dot{\bf H}^P = {\bf M}^P-{\bf v}_P\times{\bf G}.
\end{align}

::: {.callout-important title="Note!"}
$\dot{\bf H}^P$ is **not** necessarily equal to ${\bf M}^P$. The two agree only in special cases.
:::

Two important special cases:

1. $P = O$ (fixed origin, ${\bf v}_O = {\bf 0}$): $\dot{\bf H}^O = {\bf M}^O = \sum_{k=1}^n{\bf r}_k\times{\bf F}_k$.
2. $P = C$ (center of mass, ${\bf v}_C\times{\bf G} = {\bf 0}$): $\dot{\bf H}^C = {\bf M}^C = \sum_{k=1}^n({\bf r}_k-{\bf r}_C)\times{\bf F}_k$.

::: {.callout-important title="Note!"}
The conditions for conservation of angular momentum of a system are analogous to those for a single particle.
:::

*Example: A pendulum suspended from a cart on smooth horizontal rails — write the BoAM in 3 forms.*

## Work-Energy Theorem

The work-energy theorem for a system of particles is:
\begin{align}
    \dot{T} = \sum_{k=1}^n{\bf F}_k\cdot{\bf v}_k,
\end{align}
and
\begin{align}
    \dot{E} = \sum_{k=1}^n{\bf F}_{nc_k}\cdot{\bf v}_k.
\end{align}
where ${\bf F}_{nc_k}$ is the nonconservative force acting on the $k$th particle and $E$ is the total energy of the system of particles.
Integrating:
\begin{align}
    T(t_2)-T(t_1) = \sum_{k=1}^n W_{{\bf F}_k,12}, \qquad W_{{\bf F}_k,12} = \int_{t_1}^{t_2}{\bf F}_k\cdot{\bf v}_k\,dt.
\end{align}

::: {.callout-important title="Note!"}
The work of a force is the integral of the force dotted with the velocity of the **point of application** of the force.
:::

::: {.callout-warning title="Example"}
**Question:** Consider the unsymmetrical dumbbell (particle 1 of mass $m$, particle 2 of mass $3m$, massless rod) moving in the smooth $\{{\bf E}_x,{\bf E}_y\}$ plane with gravity along $-{\bf E}_z$. Is the energy conserved?

::: {.callout-note title="Answer" collapse="true"}
Drawing the FBDs of both masses: ...
:::
:::


## Summary

**Kinematics of a system of particles**

Consider a system of $n$ particles each with mass $m_i$ and with position vector ${\bf r}_i$ from the origin.
The total mass of the system of particles is $m = \sum_{i=1}^n m_i$.
The center of mass of the system of particles $C$ is defined to be located at
\begin{align}
    {\bf r}_C = \frac{\sum_{i=1}^n m_i{\bf r}_i}{m}.
\end{align}
The velocity and acceleration of the center of mass are then obtained by differentiation ${\bf r}_C$.
\begin{align}
    {\bf v}_C = \frac{\sum_{i=1}^n m_i{\bf v}_i}{m},\quad\text{and}\quad{\bf a}_C = \frac{\sum_{i=1}^n m_i{\bf a}_i}{m}.
\end{align}
The linear momentum of a system of particles is the sum of the linear momenta of its constituents which is also the linear momentum of the center of mass
\begin{align}
    {\bf G} = \sum_{i=1}^n{\bf G}_i = m{\bf v}_C.
\end{align}
The angular momentum of a system of particles about point $P$ is the sum of the angular momenta of its constituents about $P$
\begin{align}
    {\bf H}^P = \sum_{i=1}^n\lp{\bf r}_i-{\bf r}_P\rp\times m_i{\bf v}_i.
\end{align}
This expression can be rewritten as
\begin{align}
    {\bf H}^P = {\bf H} + ({\bf r}-{\bf r}_P)\times {\bf G}.
\end{align}
where the angular momentum ${\bf H}^C$ of the system about the center of mass can be written equivalently as
\begin{align}
    {\bf H} =  \sum_{i=1}^n({\bf r}_i-{\bf r}_C)\times m_k{\bf v}_i= \sum_{i=1}^n({\bf r}_i-{\bf r}_C)\times m_i({\bf v}_i-{\bf v}_C).
\end{align}
The kinetic energy of the system of particles is the sum of the kinetic energies of its constituents. This is not equal to the $\frac{1}{2}m{\bf v}_C\cdot{\bf v}_C$.
\begin{align}
    T = \sum_{i=1}^nT_i = \sum_{i=1}^n\frac{1}{2}m_i{\bf v}_i\cdot{\bf v}_i = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C + \frac{1}{2}\sum_{i=1}^nm_k({\bf v}_i-{\bf v}_C)\cdot({\bf v}_i-{\bf v}_C).
\end{align}

**Kinetics of a system of particles**

The balance of linear momentum for a system of particles is obtained by adding the balance of linear momenta of the individual particles ${\bf F}_i = m_i{\bf a}_i$
\begin{align}
    {\bf F} = \sum{\bf F}_i = \sum m_i{\bf a}_i = m{\bf a}_c.
\end{align}
Here, ${\bf F}$ is the net external force acting on the system.
The balance of angular momentum of a system of particles about any point $P$ is
\begin{align}
    \dot{\bf H}^P = {\bf M}^P-{\bf v}^P\times{\bf G}.
\end{align}
If $P$ is a fixed point $O$, then this equation simplifies to ${\bf M}^O=\dot{\bf H}^O$.
If $P$ is the center of mass $C$, then this equation simplifies to ${\bf M}^C = \dot{\bf H}^C$.
The work-energy theorem for a system of particles is the sum of the work-energy theorems of the individual particles
\begin{align}
    \sum \dot{T}_i = \sum {\bf F}_i\cdot{\bf v}_i.
\end{align}
Now that different points in the system have different displacements/velocities, it is important to note that the work of a force $i$ is obtained by dotting ${\bf F}_i$ with the velocity/displacement of its point of application.

## Exercises

*The following problems are from Set 14 – System of Particles.*

**1.** [MKB 03-168] Show that during the (very brief) collision, the linear momentum of the bullet–pendulum system is conserved. *(ans. $\theta=20.7^\circ$, $n=99.8\%$)*

**2.** [MKB 03-178] Use the integral form of the BoAM; choose the origin at the green base. *(ans. $\dot\theta=26.0$ rad/s)*

**3.** [04-008] Take an appropriate cut in the rope. *(ans. $T=58.3$ lb)*

**4.** [04-009] *(ans. $D_x=1.288$ lb left, $D_y=35.5$ lb up, $N_E=25.6$ lb up)*

**5.** [04-019] Identify a direction of linear momentum conservation; the center of mass remains stationary. *(ans. $s=\frac{(m_1+m_2)x_{10}-m_2 l}{m_0+m_1+m_2}$)*
