% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrreprt}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother





\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={MECH230 Dynamics Notes},
  pdfauthor={Theresa Honein},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{MECH230 Dynamics Notes}
\author{Theresa Honein}
\date{2026-01-01}
\begin{document}
\maketitle

\renewcommand*\contentsname{Table of contents}
{
\setcounter{tocdepth}{2}
\tableofcontents
}

\bookmarksetup{startatroot}

\chapter*{Preface}\label{preface}
\addcontentsline{toc}{chapter}{Preface}

\markboth{Preface}{Preface}

These notes are prepared for the MECH230 Dynamics course at the American
University of Beirut (AUB), Lebanon. The content is adapted primarily
from O'Reilly's Primer, with selected examples and exercises drawn from
MKB's and Hibbeler's classic textbooks.

These notes begin with simpler examples to build a strong foundation,
enabling students to confidently tackle more complex problems. They are
not comprehensive; students should refer to the course textbooks as
needed.

I am grateful for the perspectives on dynamics of Professor Noel Perkins
(University of Michigan--Ann Arbor), Professor James Casey (UC
Berkeley), and Dr.~Evan Hemingway, which I gained while TAing Berkeley's
ME104.

If you find any mistakes in these notes, please let me know by filling
out \url{https://forms.gle/McKpRhMq4JejnG8V9} form.

\part{Preliminaries}

\chapter{Active Learning}\label{active-learning}

Check out these two excellent videos of Professor Noel Perkins from the
University of Michigan at Ann Arbor explaining the active learning
techniques he uses to teach dynamics:
\href{https://www.youtube.com/watch?v=wHEys-JHeb8}{Part 1},
\href{https://www.youtube.com/watch?v=96j69u4v-wE}{Part 2}. We will be
using these techniques ourselves. The techniques include:

\begin{itemize}
\tightlist
\item
  \textbf{Having an agenda for the day.} Students can add topics they
  want to discuss too.
\item
  \textbf{Frequently checking student understanding:}

  \begin{itemize}
  \tightlist
  \item
    ``Raise your hand if you understand 90\% of this or above.''
  \end{itemize}
\item
  \textbf{Encouragement:}

  \begin{itemize}
  \tightlist
  \item
    ``Ask me a question.''
  \item
    ``Persevere. Someone else is going to have this question.''
  \item
    ``I had a great question from one of your classmates.''
  \end{itemize}
\item
  \textbf{Think → Share:}

  \begin{itemize}
  \tightlist
  \item
    ``I would like you to do the first step on your own. Take a min.''
  \item
    ``Show of hands: I did it, or I know how to do it.''
  \end{itemize}
\item
  \textbf{Think → Pair → Share:} * ``I would like you to take 5 min to
  talk to your neighbor and come up with a strategy that will solve this
  problem.''

  \begin{itemize}
  \tightlist
  \item
    ``Show of hands: my neighbor convinced me this will work.''
  \item
    ``Show of hands: I am not convinced yet.''
  \end{itemize}
\item
  \textbf{This is a test:} I am testing this.

  \begin{itemize}
  \tightlist
  \item
    Test 1
  \end{itemize}
\end{itemize}

\part{Rigid Body Dynamics}

\chapter{Kinematics of Soft Robots}\label{kinematics-of-soft-robots}

\chapter{Overview}\label{overview}

This chapter covers the geometric representation of deformable
structures.

\section{Discrete Elastic Rods}\label{discrete-elastic-rods}

Here we discuss the DER framework and its application to line contact.

\subsection{Degrees of Freedom}\label{degrees-of-freedom}

Details on the nodal coordinates\ldots{}

\section{Symbolic Formulations}\label{symbolic-formulations}

Using symbolic math to derive equations of motion.

\chapter{Rigid Body Kinetics}\label{rigid-body-kinetics}

\begin{equation}\protect\phantomsection\label{eq-relative-pos}{
\mathbf{r}_{P/O} = \mathbf{r}_{G/O} + \mathbf{r}_{P/G}
}\end{equation}

The relative position vector is defined in
Equation~\ref{eq-relative-pos}.

\chapter{Work Energy Theorem}\label{work-energy-theorem}

The Koenig decomposition for the kinetic energy of a rigid body is:
\begin{align}
    T = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}{\bf H}\cdot\bomega.
\end{align} In this class, \(\bomega = \omega {\bf E}_z\),
\(\frac{1}{2}{\bf H}\cdot\bomega = \frac{1}{2}I_{zz} w^2\).

In the special case of a fixed point rotation about point \(O\), the
kinetic energy becomes \begin{align}
    T = \frac{1}{2}{\bf H}^O\cdot\bomega.
\end{align} which simplifies to \(T = \frac{1}{2}I^O_{zz} w^2\) when
\(\bomega = \omega{\bf E}_z\).

The work-energy theorem for a rigid body is: \begin{align}
    \frac{dT}{dt} = {\bf F}\cdot{\bf v}_C+{\bf M}\cdot\bomega = \sum_{i=1}^N{\bf F}_i\cdot{\bf v}_i+{\bf M}_e\cdot\bomega.
\end{align} Integrating the above expression with respect time between
\(t_A\) and \(t_B\), we obtain \begin{align}
    \begin{split}
        T_B-T_A = W_{{\bf F},AB}+W_{{\bf M},AB} = \sum_{i=1}^N W_{{\bf F}_i,AB}+W_{{\bf M}_e,AB},
        \label{eq:work-energy-theorem-RB}
    \end{split}
\end{align} where the work of the forces and moments is calculates as
follows: \begin{align}
    \begin{split}
        W_{{\bf F},AB} &= \int_{t_A}^{t_B}{\bf F}\cdot{\bf v}_Cdt,\\
        W_{{\bf M},AB} &= \int_{t_A}^{t_B}{\bf M}^C\cdot\bomega dt,\\
        W_{{\bf F}_i,AB} &= \int_{t_A}^{t_B}{\bf F}_i\cdot{\bf v}_i dt,\\
        W_{{\bf M}_e,AB} &= \int_{t_A}^{t_B}{\bf M}_e\cdot\bomega dt,\\
    \end{split}
\end{align} Notice that in calculating the work of a force, we dot the
force with the velocity of its point of application.

As in particles and systems of particles, this theorem can be used to
establish conservation of the total energy of a rigid body during a
motion.

In the following section, we derive these results.

\section{Koenig's Decomposition}

By definition, the kinetic energy \(T\) of a rigid body is the
(continuous) sum of the kinetic energies of the differential masses
composing it: \begin{align}
    T = \frac{1}{2}\int_{\mathcal{R}}{\bf v}\cdot{\bf v}\rho dv.
\end{align} We can introduce the \({\bf v}_C\) term through
\begin{align}
    \begin{split}
        & {\bf v} = {\bf v}_C+\bomega\times\bpi,\\
        & \bpi = {\bf r}-{\bf r}_C,\\
        & \bomega = \omega_x{\bf e}_x+\omega_y{\bf e}_y+\omega_z{\bf e}_z.
    \end{split}
\end{align} Substituting the above expressions in the kinetic energy
integral, we get \begin{align}
    T = \frac{1}{2}\int_{\mathcal{B}}\lp{\bf v}_C\cdot{\bf v}_C+2{\bf v}_C\cdot(\bomega\times\bpi)+(\bomega\times\bpi)\cdot(\bomega\times\bpi)  \rp dm
\end{align} For the first two terms in the above integral, \begin{align}
    \begin{split}
        & \frac{1}{2}\int_{\mathcal{B}}{\bf v}_C\cdot{\bf v}_C dm = \frac{{\bf v}_C\cdot{\bf v}_C}{2}\int dm = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C.\\
        & \int_{\mathcal{B}}{\bf v}_C\cdot(\bomega\times\bpi)dm = {\bf v}_C\cdot\lp\bomega\times\int\pi dm \rp = 0.
    \end{split}
\end{align} Thus, \begin{align}
    T = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}\int_{\mathcal{B}}(\bomega\times\bpi)\cdot(\bomega\times\bpi)dm
\end{align} we can simplify\footnote{Here we used 
\begin{align*}
    ({\bf a}\times{\bf b})\cdot({\bf c}\times{\bf d}) = ({\bf a}\cdot{\bf c})({\bf b}\cdot{\bf d})-({\bf a}\cdot{\bf d})({\bf b}\cdot{\bf c}).
\end{align*}} \begin{align}
    (\bomega\times\bpi)\cdot(\bomega\times\bpi) = ((\bpi\cdot\bpi)\bomega-(\bpi\cdot\bomega)\cdot\bpi).
\end{align} Recall that \begin{align}
    {\bf H}^C = \int_{\mathcal{B}}\bpi\times(\omega\times\bpi)dm = \int_{\mathcal{B}}((\bpi\cdot\bpi)\bomega-(\bpi\cdot\bomega)\bpi)dm.
\end{align} Hence, we obtain the Koenig decomposition: \begin{align}
    T = \frac{1}{2}m{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}{\bf H}\cdot\bomega.
\end{align}

\subsection{Fixed Point Rotation}

In the case of fixed point rotation,
\({\bf v}_C = \bomega\times({\bf r}_C-{\bf r}_O)\) and
\({\bf v}_O = 0\),
then\footnote{Here we use the cyclical property of the scalar triple product: $${\bf a}\cdot({\bf b}\times{\bf c}) = {\bf b}\cdot({\bf c}\times{\bf a}) = {\bf c}\times({\bf a}\times{\bf b}).$$}
\begin{align}
    \begin{split}
        T &= \frac{1}{2}{\bf G}\cdot(\bomega\times{\bf r}_{C/O})+\frac{1}{2}{\bf H}\cdot\bomega\\
        &= \frac{1}{2}\bomega\cdot({\bf r}_{C/O}\times{\bf G})+\frac{1}{2}{\bf H}^C\cdot\bomega\\
        &= \frac{1}{2}({\bf r}_{C/O}\times{\bf G}+{\bf H})\cdot\bomega\\
        &= \frac{1}{2}{\bf H}^O\cdot\bomega.
    \end{split}
\end{align}

\section{The Work-Energy Theorem}

\begin{align}
    \dot{T} = \frac{1}{2}m\dot{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}m{\bf v}_C\cdot\dot{\bf v}_C+\frac{1}{2}\dot{\bf H}\cdot\bomega+\frac{1}{2}{\bf H}\cdot\dot{\bomega}.
\end{align} We need to show that
\(\dot{\bf H}\cdot\bomega = {\bf H}\cdot\dot{\bomega}\). \begin{align}
    \begin{split}
        \balpha = \dot{\bomega} &= \frac{d}{dt}(\omega_x{\bf e}_x+\omega_y{\bf e}_y+\omega_z{\bf e}_z)\\
        &= \dot{\omega}_x{\bf e}_x+\dot{\omega}_y{\bf e}_y+\dot{\omega}_z{\bf e}_z+\omega_x\dot{\bf e}_x+\omega_y\dot{\bf e}_y+\omega_z\dot{\bf e}_z\\
        &= \dot{\omega}_x{\bf e}_x+\dot{\omega}_y{\bf e}_y+\dot{\omega}_z{\bf e}_z+\bomega\times(\omega_x{\bf e}_x+\omega_y{\bf e}_y+\omega_z{\bf e}_z)\\
        &= \dot{\omega}_x{\bf e}_x+\dot{\omega}_y{\bf e}_y+\dot{\omega}_z{\bf e}_z.
    \end{split}
\end{align}

Another direct calculation using this expression for \(\balpha\) shows
that \begin{align}
    \begin{split}
        {\bf H}^C\cdot\dot{\bomega} = (I_{xx}\omega_x+I_{xy}\omega_y+I_{xz}\omega_z)\dot{\omega}_x+(I_{xy}\omega_x+I_{yy}\omega_y+I_{yz}\omega_z)\dot{\omega}_y+(I_{xz}\omega_x+I_{yz}\omega_y+I_{zz}\omega_z)\dot{\omega}_z.
    \end{split}
\end{align}

Comparing this to the corresponding expression for
\(\dot{H}^C\cdot\bomega\), we find that they are equal. Consequently,
\begin{align}
    \dot{T} = \frac{1}{2}m\dot{\bf v}_C\cdot{\bf v}_C+\frac{1}{2}m{\bf v}_C\cdot\dot{\bf v}_C+\frac{1}{2}\dot{\bf H}\cdot\bomega+\frac{1}{2}\dot{\bf H}\cdot\bomega.
\end{align} This result implies that \begin{align}
    \dot{T} = m\dot{\bf v}_C\cdot{\bf v}_C+\dot{\bf H}\cdot\bomega.
\end{align} Invoking the balance of linear momentum and the balance of
angular momentum, we obtain the work-energy theorem: \begin{align}
    \dot{T} = {\bf F}\cdot{\bf v}_C+{\bf M}\cdot\bomega.
\end{align} You should notice how this is a natural extension of the
work-energy theorem for a single particle. We can then integrate these
expressions of the work-energy theorem to get equation
(\ref{eq:work-energy-theorem-RB}).

\section{An alternative form of the work-energy theorem}

Recall, \begin{align}
    \begin{split}
        & {\bf F} = \sum_{i=1}^K{\bf F}_i\\
        & {\bf M}^C = {\bf M}_e+\sum_{i=1}^K({\bf r}_i-{\bf r}_C)\times{\bf F}_i
    \end{split}
\end{align} Hence, the mechanical power of the resultant forces and
moments can be written as \begin{align}
    {\bf F}\cdot{\bf v}_C+{\bf M}^C\cdot\bomega = \lp\sum_{i=1}^K{\bf F}_i\rp\cdot{\bf v}_C+{\bf M}_e\cdot\bomega+\lp \sum_{i=1}^K({\bf x}_i-{\bf x}_C)\times{\bf F}_i\rp\cdot\bomega
\end{align} With some minor manipulations involving the identity
\({\bf a}\cdot({\bf b}\times{\bf c}) = {\bf c}\cdot({\bf a}\times{\bf b})\)
and noting that
\({\bf v}_i = {\bf v}_C+\bomega\times({\bf x}_i-{\bf x}_C)\), we find
that \begin{align}
    {\bf F}\cdot{\bf v}_C+{\bf M}\cdot\bomega = \sum_{i=1}^K{\bf F}_i\cdot{\bf v}_i+{\bf M}_e\cdot\bomega.
\end{align} In conclusion, we have an alternative form of the
work-energy theorem that proves to be far easier to use in applications:
\begin{align}
    \dot{T} = \sum_{i=1}^K{\bf F}_i\cdot{\bf v}_i+{\bf M}_e\cdot\bomega.
\end{align}

\subsection{Example: A bar of length $\ell$ and mass $m$ pendulum fixed at its end}

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \draw[thick] (0,0) -- (3,-4);
        \fill[black] (0,0) circle (0.1);
        \draw[thick,->] (0,0) -- (0,-1) node[below] {${\bf E}_x$};
        \draw[thick,->] (0,0) -- (1,0) node[right] {${\bf E}_y$};
        \draw[thick,->,blue] (0,0) -- (0.6,-0.8) node[below] {${\bf e}_x$};
        \draw[thick,->,blue] (0,0) -- (0.8,0.6) node[above] {${\bf e}_y$};
    \end{tikzpicture}
\end{figure}

\begin{align}
    \begin{split}
        & U = -\frac{mg\ell}{2}\cos(\theta),\\
        & E = \frac{m\ell^2}{6}\dot{\theta}^2-\frac{mg\ell}{2}\cos(\theta) = const.
    \end{split}
\end{align}

\subsection{Example: What happens if you attach a dead mass at the end of the rod?}

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \draw[thick] (0,0) -- (3,-4);
        \fill[black] (0,0) circle (0.1);
        \draw[thick,->] (0,0) -- (0,-1) node[below] {${\bf E}_x$};
        \draw[thick,->] (0,0) -- (1,0) node[right] {${\bf E}_y$};
        \draw[thick,->,blue] (0,0) -- (0.6,-0.8) node[below] {${\bf e}_x$};
        \draw[thick,->,blue] (0,0) -- (0.8,0.6) node[above] {${\bf e}_y$};
        \fill[black] (3,-4) circle (0.4);
    \end{tikzpicture}
\end{figure}

\begin{align}
    & U = -\frac{mg\ell}{2}\cos(\theta)-m_e g\ell\cos(\theta)
\end{align}

\underline{What happens if there is a follower force applied at the end of the rod always perpendicular to it?}

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \draw[thick] (0,0) -- (3,-4);
        \fill[black] (0,0) circle (0.1);
        \draw[thick,->] (0,0) -- (0,-1) node[below] {${\bf E}_x$};
        \draw[thick,->] (0,0) -- (1,0) node[right] {${\bf E}_y$};
        \draw[thick,->,blue] (0,0) -- (0.6,-0.8) node[below] {${\bf e}_x$};
        \draw[thick,->,blue] (0,0) -- (0.8,0.6) node[above] {${\bf e}_y$};
        
        \draw[thick,->] (3,-4) -- (4.6,-2.8) node[above] {${\bf P}$};
    \end{tikzpicture}
\end{figure}

\({\bf P}\) is actively changing depending on what the configuration is.

This is a nonconservative system.

\section{Conservative Moments}

Moments can also be conservative.

A torsional spring is just a model for some kind of elasticity at the
joint.

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        % Parameters for the spiral
        \draw[domain=0:1000, samples=100, smooth, variable=\t] 
            plot ({0.0005*\t * cos(\t)}, {0.0005*\t * sin(\t)});
        \draw[thick] (0,0) -- (3,-4);
        \fill[black] (0,0) circle (0.1);
        \draw[thick,->] (0,0) -- (0,-1) node[below] {${\bf E}_x$};
        \draw[thick,->] (0,0) -- (1,0) node[right] {${\bf E}_y$};
        \draw[thick,->,blue] (0,0) -- (0.6,-0.8) node[below] {${\bf e}_x$};
        \draw[thick,->,blue] (0,0) -- (0.8,0.6) node[above] {${\bf e}_y$};
    \end{tikzpicture}
\end{figure}

The torsional spring with stiffness \(K\) creates a couple \begin{align}
    {\bf M}_S = K\theta(-{\bf E}_Z)
\end{align} The power of the couple is \begin{align}
    {\bf M}_S\cdot\bomega = -K\theta\dot{\theta} = -\frac{d}{dt}\lp K\frac{\dot{\theta}^2}{2} \rp.
\end{align} Hence, \(U = \frac{1}{2}K\dot{\theta}^2\) for the torsional
spring.

A constant couple \({\bf M}_e = M_e{\bf E}_Z\) is also conservative. Its
power is \begin{align}
    {\bf M}_e\cdot\bomega = M_e\dot{\theta} = -\frac{d}{dt}(-M_e\theta).
\end{align} Hence, \(U = -M_e\theta\) for a constant couple.

Note. This only works for \(\bomega = \omega{\bf E}_z\). Constant
couples are not conservative in general.

\newpage
\section{Impact Problems / Center of Percussion}

For rigid bodies, impact ideas are the same as in systems of particles,

\underline{Example}

Consider a projectile fired at a pendulum in the horizontal plane (no
gravity). What is \(x\) so that the reaction for at the pin is almost
zero during impact.

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \draw[thick] (0,0) -- (0,-4);
        \fill (0,0) circle (0.1);
        \fill (-2,-3) circle (0.25);
        \draw[thick,->] (-2,-3) -- (-1,-3) node[above] {${\bf v}_0$};
        \draw[->] (1,0) -- (1,-3) node[midway, right] {$x$};
    \end{tikzpicture}
\end{figure}

If \(x\) is big enough, the horizontal component of reaction force at
\(O\) will point to the left. If \(x\) is small enough, the horizontal
component of the reaction force at \(O\) will point to the right.
Somewhere in between, the reaction force at \(O\) will be minimized.
This optimal location \(x\) is called the center of percussion.

\bookmarksetup{startatroot}

\chapter{Summary}\label{summary}

In summary, this book has no content whatsoever.

\bookmarksetup{startatroot}

\chapter*{References}\label{references}
\addcontentsline{toc}{chapter}{References}

\markboth{References}{References}

\protect\phantomsection\label{refs}




\end{document}
