---
title: "Basic Particle Kinematics Concepts"
---

## Position, Velocity, and Acceleration

Consider a particle moving in $\mathbb{E}^3$. Its position vector relative to a fixed origin $O$ is denoted by ${\bf r} \in \mathbb{E}^3$. As time $t$ evolves, so does ${\bf r}(t)$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}

    % Draw the random smooth curve
    \draw[thick] plot[smooth, tension=1] coordinates {(-2,-1) (0,2) (2,-1) (4,1)};

    % Draw the particle on the curve
    \node[circle, fill=red, inner sep=2pt, label=below:{$P$}] at (0,2) {};

    % Draw the origin
    \node[circle, fill=black, inner sep=2pt, label=below:{$O$}] at (-2,-2) {};

    % Draw the position vector to the particle
    \draw[->, thick, blue] (-2,-2) -- (0,2) node[midway, above left] {${\bf r}$};

\end{tikzpicture}
```

Let $\Delta t$ be a finite time elapsed:
\begin{align}
    \Delta{\bf r} = {\bf r}(t+\Delta t)-{\bf r}(t).
\end{align}
As $\Delta t \rightarrow 0$, $\Delta t$ becomes $dt$ and $\Delta {\bf r}$ becomes $d{\bf r}$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\usetikzlibrary{calc}
\begin{tikzpicture}

% --- Left Diagram ---
\draw[dashed] (-0.5,1.8) to[out=0,in=160] (2,1) to[out=-20,in=140] (4,0.5);

\draw[thick] (0,0) -- (0,1);
\draw[thick] (0,0) -- (-0.7,-0.5);
\draw[thick] (0,0) -- (0.7,-0.5);

\coordinate (A) at (1.5,1.2);
\coordinate (B) at (3,0.8);

\fill (A) circle (0.05);
\fill (B) circle (0.05);

\draw[->,thick] (0,0) -- (A);
\draw[->,thick] (0,0) -- (B);
\draw[->,thick] (A) -- (B);

\node[above] at (A) {$\mathbf{r}(t)$};
\node[right] at (B) {$\mathbf{r}(t+\Delta t)$};
\node[above] at ($(A)!0.5!(B)$) {$\Delta \mathbf{r}$};

% --- Right Diagram ---
\begin{scope}[xshift=7cm]
\draw[dashed] (-0.5,1.8) to[out=20,in=140] (4,0.5);

\node at (0,-0.3) {$O$};

\coordinate (C) at (1.8,1.9);
\coordinate (D) at (2,1.8);
\draw[->,thick] (0,0) -- (C);
\draw[->,thick] (0,0) -- (D);
\draw[->,thick] (C) -- ++(1.2,-0.5) node[right] {$\mathbf{v}$};

\node at (2,3) {zoom in};
\end{scope}

\end{tikzpicture}
```

The (absolute) velocity vector ${\bf v}$ of the particle is
\begin{align}
    {\bf v} = \frac{d{\bf r}}{dt}=\dot{\bf r} = \lim_{\Delta t \rightarrow 0}\frac{{\bf r}(t+\Delta t)-{\bf r}(t)}{\Delta t}.
\end{align}

::: {.callout-tip title="Think!"}
**Question:** What is the direction of the velocity vector?

**Answer:** ${\bf v}$ is always tangent to the path.
:::

The speed of the particle is the magnitude of the velocity vector $v = \lnorm{\bf v}\rnorm$.

The (absolute) acceleration is
\begin{align}
    {\bf a} = \frac{d{\bf v}}{dt} = \dot{\bf v} = \ddot{\bf r} = \frac{d^2{\bf r}}{dt^2}.
\end{align}

## Arc-Length and the Unit Tangent Vector

Define the arc-length parameter $s$ such that
\begin{align}
    \frac{ds}{dt} = \lnorm {\bf v}\rnorm.
\end{align}

$\frac{ds}{dt}$ is the speed of the particle. Recall that ${\bf v} = \frac{d{\bf r}}{dt}$, so
\begin{align}
    \frac{ds}{dt} = \lnorm\frac{d{\bf r}}{dt}\rnorm.
\end{align}

So $ds$ is the length of the differential $d{\bf r}$ along the curve. Integrating:
\begin{align}
    s(t)-s_0 = \int_{t_0}^t\frac{ds}{dt}(\tau)d\tau = \int_{t_0}^t\sqrt{{\bf v}(\tau)\cdot{\bf v}(\tau)}\,d\tau.
\end{align}

Notes:

- $s(t)-s_0$ is the distance traveled by the particle along its path $\mathcal{C}$ during the time interval $[t_0, t]$.
- $s(t_0)=s_0$ where $t_0$ and $s_0$ are initial conditions.
- The dummy variable $\tau$ is used (as opposed to $t$) when evaluating this integral because we are integrating the speed as it varies between $t_0$ and $t$.

::: {.callout-tip title="Think!"}
**Question:** Under what conditions is $s(t)$ invertible so that $t(s)$ is well defined?

**Answer:** If the function $s(t)$ is one-to-one, it can be inverted, and time can be written in terms of arc-length $t = t(s)$.
:::

::: {.callout-tip title="Think!"}
**Question:** Why is
\begin{align}
    s(t)-s_0 \neq \lnorm {\bf r}(t)-{\bf r}_0\rnorm?
\end{align}
Under what conditions does the equality hold?

**Answer:** The equality holds under rectilinear motion or over an infinitesimal time interval.
:::

We define the unit tangent vector ${\bf e}_t$:
\begin{align}
    \frac{\bf v}{\lnorm{\bf v}\rnorm} = {\bf e}_t, \qquad {\bf v} = \frac{d{\bf r}}{dt} = \frac{d{\bf r}}{ds}\frac{ds}{dt} = \dot{s}{\bf e}_t.
\end{align}

::: {.callout-tip title="Think!"}
**Question:** Verify that ${\bf e}_t$ is a unit vector.

**Answer:** …
:::

Note the relationships
\begin{align*}
    {\bf v} &= \frac{d{\bf r}(t)}{dt} = \frac{d{\bf r}}{ds}\frac{ds}{dt},\\
    {\bf a} &= \frac{d{\bf v}(t)}{dt} = \frac{d^2{\bf r}}{ds^2}\lp\frac{ds}{dt}\rp^2+\frac{d{\bf r}}{ds}\frac{d^2 s}{dt^2}.
\end{align*}

Notice: the acceleration vector is due to changes in both the magnitude and direction of the velocity vector.

::: {.callout-tip title="Think!"}
**Question:** Is the unit vector ${\bf e}_t$ constant in general? Under what conditions is it constant?

**Answer:** No. It is constant only during rectilinear motion in a fixed direction.
:::

### Example

Consider a particle moving in space with position vector
\begin{align}
    {\bf r}(t) = R_0\lp\cos(\omega t){\bf E}_x+\sin(\omega t){\bf E}_y\rp.
\end{align}

Calculate and plot ${\bf v}$, ${\bf a}$, ${\bf v}\times {\bf a}$, $\lnorm{\bf r}\rnorm$. Show that the motion satisfies:
\begin{align}
    \ddot{x}+\omega^2 x=0, \qquad \ddot{y}+\omega^2 y=0.
\end{align}

::: {.callout-warning title="Example"}
**Question:** Consider a particle $P$ moving in $\mathbb{E}^2$ with position vector
\begin{align*}
    {\bf r} = R_0\lp\cos\lp\omega t\rp{\bf E}_x+\sin\lp\omega t\rp{\bf E}_y\rp.
\end{align*}
Calculate ${\bf r}\times{\bf a}$ and ${\bf r}\cdot{\bf a}$. Describe the motion of the particle.

**Answer:** The velocity and acceleration are:
\begin{align*}
    {\bf v} &= R_0\omega\lp-\sin\lp\omega t\rp{\bf E}_x+\cos\lp\omega t\rp{\bf E}_y\rp,\\
    {\bf a} &= -R_0\omega^2\lp\cos\lp\omega t\rp{\bf E}_x+\sin\lp\omega t\rp{\bf E}_y\rp.
\end{align*}
Note that $\dot{\bf E}_x={\bf 0}$ and $\dot{\bf E}_y={\bf 0}$ since the basis is fixed, and the chain rule applies.

Noting that ${\bf a} = -\omega^2{\bf r}$:
\begin{align*}
    {\bf r}\times{\bf a} = {\bf r}\times(-\omega^2{\bf r}) = -\omega^2{\bf r}\times{\bf r} = {\bf 0}.
\end{align*}

For ${\bf r}\cdot{\bf a}$:
\begin{align*}
    {\bf r}\cdot{\bf a} &= -R_0^2\omega^2\lp\cos^2(\omega t)+\sin^2(\omega t)\rp = -R_0^2\omega^2.
\end{align*}
:::

## Cartesian Coordinates

$\{{\bf E}_x,{\bf E}_y,{\bf E}_z\}$ is a basis for $\mathbb{E}^3$. Since this basis is fixed, we have:
\begin{align}
    {\bf r} &= x{\bf E}_x+y{\bf E}_y+z{\bf E}_z,\\
    {\bf v} &= \dot{x}{\bf E}_x+\dot{y}{\bf E}_y+\dot{z}{\bf E}_z,\\
    {\bf a} &= \ddot{x}{\bf E}_x+\ddot{y}{\bf E}_y+\ddot{z}{\bf E}_z.
\end{align}

Later we will also use the cylindrical polar basis $\{{\bf e}_r,{\bf e}_\theta,{\bf E}_z\}$ and the Serret-Frenet triad $\{{\bf e}_t,{\bf e}_n,{\bf e}_b\}$.

## Rectilinear Motion

- *rectus* = straight, *linea* = line

Due to constraints or initial conditions, the body moves on a straight line, e.g. a car on a straight road or a ball tossed vertically.

We take ${\bf E}_x$ parallel to the line and ${\bf c}$ to be a constant vector. Then:
\begin{align}
    {\bf r}(t) &= x(t){\bf E}_x+{\bf c},\\
    {\bf v}(t) &= \dot{x}{\bf E}_x = v(t){\bf E}_x,\\
    {\bf a}(t) &= \ddot{x}{\bf E}_x = a(t){\bf E}_x.
\end{align}

$\frac{ds}{dt}=\left|\frac{dx}{dt}\right|$, so unless $\dot{x}>0$ or $\dot{x}<0$ throughout, $x$ and $s$ cannot be easily interchanged.

### Identities for Rectilinear Motion

Recall $v = \frac{dx}{dt}$ and $a = \frac{dv}{dt}$. Applying the chain rule:
\begin{align}
    a = \frac{dv}{dt} = \frac{dv}{dx}\frac{dx}{dt} = v\frac{dv}{dx}.
\end{align}

Three useful identities:
\begin{align}
    ds &= v\, dt, \qquad dv = a\, dt, \qquad v\,dv = a\,ds.
\end{align}

### Given Acceleration as a Function of Time

From $dv = a\, dt$:
\begin{align}
    v(t)-v(t_0) = \int_{t_0}^{t}a(\tilde{t})\,d\tilde{t}.
\end{align}
From $ds = v\, dt$:
\begin{align}
    s(t_1)-s(t_0) = \int_{t_0}^{t_1}v(\tilde{t})\,d\tilde{t}.
\end{align}

### Given Acceleration as a Function of Speed

From $dv = a\, dt$ with $a = a(v)$:
\begin{align}
    t(v) - t(v_0) = \int_{v_0}^{v}\frac{d\tilde{v}}{a(\tilde{v})}.
\end{align}
Using $a = v\frac{dv}{ds}$:
\begin{align}
    s(t)-s_0 = \int_{v_0}^{v}\frac{\tilde{v}\,d\tilde{v}}{a(\tilde{v})}.
\end{align}

### Given Acceleration as a Function of Displacement

Using $v\,dv = a\,ds$ with $a = a(s)$:
\begin{align}
    \frac{1}{2}\lp v^2(s)-v^2(s_0)\rp = \int_{s_0}^{s}a(\tilde{s})\,d\tilde{s}.
\end{align}

#### Constant Acceleration

**Example: Gravity.** Throw a ball up; take the hand as origin. Here $a = -g$.

\begin{align}
    v(t) &= v_0-gt,\\
    s(t) &= s_0+v_0t-\frac{g}{2}t^2.
\end{align}

**Example: Vehicle Braking.** Using $v\,dv = a\,ds$:
\begin{align}
    v^2(s)-v^2(s_0) = 2a\,\Delta s.
\end{align}
