---
title: "Friction Force"
---

Most theories of friction arise from studies by the French scientist Charles Augustin de Coulomb (1736–1806).

## Introduction

Consider a block on a rough horizontal surface pushed with force ${\bf P} = P{\bf E}_x$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick] (0,0) rectangle (3,1.5);
    \draw (-1,0) -- (6,0);
    \draw[->] (3,0.75) -- (5,0.75) node[right] {$\mathbf{P}$};
    \draw[->] (-2,0) -- (-1,0) node[right, above] {${\bf E}_x$};
    \draw[->] (-2,0) -- (-2,1) node[above] {${\bf E}_y$};
\end{tikzpicture}
```

Upon increasing $P$, we observe:

- For small $P$, the block remains at rest.
- Beyond a critical value $P = P^*$, the block starts to move.
- Once in motion, a constant value $P = P^{**}$ is required to maintain constant speed.
- Both $P^*$ and $P^{**}$ are proportional to the normal force $\lnorm{\bf N}\rnorm$.

## Formalism: Coulomb's Theory of Friction

We distinguish two types of friction: static and kinetic.

### Static Friction

When the relative velocity is zero (${\bf v}_{rel} = {\bf 0}$), the friction force ${\bf F}_f$ is **unknown** and lies along the contact surface. We must solve for both its magnitude and direction. We say the static friction force points in the direction that **impedes motion**.

::: {.callout-tip title="Think!"}
**Question:** Consider a block on a rough inclined surface that is not moving. Draw the FBD and write the equation of motion. Do you have as many equations as unknowns?

::: {.callout-note title="Answer" collapse="true"}
Yes. The friction and normal forces have unknown magnitudes, but the block's acceleration is known (it matches the surface acceleration). There are as many equations as unknowns.
:::
:::

The static friction criterion:
\begin{align}
    \lnorm{\bf F}_f\rnorm \leq \mu_s\lnorm{\bf N}\rnorm,
\end{align}
where $\mu_s$ is the static friction coefficient.

### Kinetic Friction

When two bodies are in relative motion (${\bf v}_{rel} \neq {\bf 0}$), the friction force is **known**:
\begin{align}
    {\bf F}_f = \mu_k\lnorm{\bf N}\rnorm\lp-\frac{{\bf v}_{rel}}{\lnorm{\bf v}_{rel}\rnorm}\rp,
\end{align}
where $\mu_k$ is the kinetic friction coefficient.

Remarks:

- $\mu_k$ is also written $\mu_d$ (dynamic friction coefficient).
- Kinetic friction is a prescription, not a constraint force.
- The magnitude of kinetic friction does not depend on speed.

### Example: Crate Pushed with Incremental Force

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick] (0,0) rectangle (3,1.5);
    \draw (-1,0) -- (6,0);
    \draw[->] (3,0.75) -- (5,0.75) node[right] {$\mathbf{P}$};
    \draw[->] (-2,0) -- (-1,0) node[right, above] {${\bf E}_x$};
    \draw[->] (-2,0) -- (-2,1) node[above] {${\bf E}_y$};
\end{tikzpicture}
```

1. Kinematics: ${\bf r} = x{\bf E}_x+y_0{\bf E}_y$, ${\bf v} = \dot{x}{\bf E}_x$, ${\bf a} = \ddot{x}{\bf E}_x$.

2. FBD with forces ${\bf P} = P{\bf E}_x$, ${\bf W} = -mg{\bf E}_y$, ${\bf N} = N{\bf E}_y$, ${\bf F}_f = F_f(-{\bf E}_x)$.

   When kinetic: ${\bf F}_f = -\mu_k\lnorm{\bf N}\rnorm\,\text{sgn}(\dot{x}){\bf E}_x$.

3. BoLM:
\begin{align}
    P{\bf E}_x-mg{\bf E}_y+N{\bf E}_y-F_f{\bf E}_x = m\ddot{x}{\bf E}_x.
\end{align}

4. Initially at rest (static friction). Setting $\ddot{x}=0$:
\begin{align}
    P = F_f, \quad N = mg.
\end{align}
The block stays at rest while $P \leq \mu_s mg$. The maximum static value is $P^* = \mu_s mg$.

Once sliding, ${\bf F}_f = -\mu_k mg{\bf E}_x$ and:
\begin{align}
    \ddot{x} = \frac{P}{m}-\mu_k g.
\end{align}
For constant speed, $P^{**} = \mu_k mg$.

### Example: A Box on a Ramp

*(MKB problem 03/002.)*

When it is unknown whether the block is in motion:

1. Assume sticking: ${\bf v}_{rel} = {\bf 0}$.
2. Calculate the static friction force.
3. Check the static friction criterion.
   - If satisfied: the block sticks. Analysis complete.
   - If not satisfied: the block slides. Apply kinetic friction and resolve.

## Relative Velocity

A common misconception: if an object is moving, the friction acting on it must be kinetic. Counter-example: a crate on a moving truck. If the crate moves with the truck (same velocity), the friction between them is **static**. Only if the crate slides on the truck bed is the friction kinetic.

Similar examples include a person in an elevator and a person riding a bus.

### Example: Nonzero Relative Velocity

- **Box held fixed on a moving conveyor belt:** The belt moves, so ${\bf v}_{rel} = -{\bf v}_c \neq {\bf 0}$ — friction is kinetic. The box is stationary in space but moving relative to the belt.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick] (0,0) rectangle (3,1.5);
    \draw (-1,0) -- (6,0);
    \draw[->] (0,0.75) -- (-2,0.75) node[left] {$\mathbf{P}$};
    \draw[->] (6.2,0) -- (7.5,0) node[right] {${\bf v}_c$};
    \draw[->] (-5,0) -- (-4,0) node[right, above] {${\bf E}_x$};
    \draw[->] (-5,0) -- (-5,1) node[above] {${\bf E}_y$};
\end{tikzpicture}
```

- **Person standing in an accelerating bus:** Friction is static. The impending motion of the person (relative to the bus) is toward the rear, so the static friction force points toward the front.

- **Car rolling without slipping:** Friction is static and points in the direction of motion, opposing the tendency of the tire to slip backward.

## Normal and Friction Force Prescription

See [this video](https://www.youtube.com/watch?v=XNdP7Nk850s).

- For a particle moving on a **space curve**: ${\bf N} = N_n{\bf e}_n+N_b{\bf e}_b$ and ${\bf F}_f = F_f{\bf e}_t$.
- For a particle moving on a **surface** with normal ${\bf n}$ and tangent directions ${\bf t}_1$, ${\bf t}_2$: ${\bf N} = N{\bf n}$ and ${\bf F}_f = F_{f1}{\bf t}_1+F_{f2}{\bf t}_2$.


## Summary

- **Static friction**: $\|\mathbf{F}_f\|\le\mu_s\|\mathbf{N}\|$.
- **Kinetic (Coulomb) friction**: $\mathbf{F}_f = -\mu_k\|\mathbf{N}\|\dfrac{\mathbf{v}_{\mathrm{rel}}}{\|\mathbf{v}_{\mathrm{rel}}\|}$.

Strategy: assume sticking; check the static criterion. If violated, resolve with kinetic friction.

## Lecture Videos

{{< video https://youtu.be/_fvCFtpL3c8 title="Explaining the friction force" width="70%" >}}


## Exercises

*The following problems are from Set 10 – Friction.*

**1.** [MKB 03-002] Consider taking $\mathbf{E}_x$ along and $\mathbf{E}_y$ perpendicular to the incline. Assume sticking first; check the static criterion. *(ans. (a) $a=1.118$ m/s$^2$ down incline; (b) $a=0$; (c) $a=2.04$ m/s$^2$ up incline)*

**2.** [MKB 03-030] Apply the 4 steps twice: first for the combined system, then for mass $m_2$ alone. *(ans. $0.0577(m_1+m_2)g \le P \le 0.745(m_1+m_2)g$)*

**3.** [MKB 03-047] Compare the Serret–Frenet basis here with Problem 03-043 from Set 07. *(ans. $N=156.2$ lb, $a=27.7$ ft/sec$^2$)*

**4.** [03-049] Set up a cylindrical polar coordinate system with origin at the base of the cylinder axis; apply the 4 steps to the small object $A$. *(ans. $\omega=\sqrt{\mu_s g/r}$)*

**5.** [03-050] This determines the maximum lateral acceleration before slipping. *(ans. $a_n=0.818g$, $F=2460$ lb)*

**6.** [03-071] Polar coordinate system with origin at the centre of the disk. *(ans. $N=\frac{\mu_s g}{4\pi^2}\left(\frac{1}{r\alpha}-1\right)$)*

{{< video https://youtu.be/RDFfE0tFTnk title="Solution – Problem 03/071" width="70%" >}}

