---
title: "Gravitational Force Model"
---

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 2
\begin{tikzpicture}

\coordinate (O) at (0,0);
\coordinate (m1) at (0,1.5);
\coordinate (m2) at (2.5,1);

\fill (m1) circle (2pt) node[above] {$m_1$};
\fill (m2) circle (2pt) node[above] {$m_2$};
\node[left] at (O) {$O$};

\draw[->,thick] (O) -- (m1) node[midway,left] {${\bf r}_1$};
\draw[->,thick] (O) -- (m2) node[midway,below] {${\bf r}_2$};
\draw[->,thick] (m1) -- (m2) node[midway,above] {${\bf r}_2 - {\bf r}_1$};

\end{tikzpicture}
```

Call ${\bf r}_2 - {\bf r}_1 = {\bf r}_{2/1}$. According to Newton's law of gravitation, the force on $m_1$ due to $m_2$ is:
\begin{align}
    {\bf F}_1 = G \frac{m_1 m_2}{\lnorm{\bf r}_{2/1}\rnorm^2}\frac{{\bf r}_{2/1}}{\lnorm{\bf r}_{2/1}\rnorm} = G\frac{m_1 m_2}{\lnorm{\bf r}_{2/1}\rnorm^3}{\bf r}_{2/1}.
\end{align}
where $G = 6.67408\times10^{-11}$ m$^3$ kg$^{-1}$ s$^{-2}$ is the gravitational constant.


## Example: Attraction Between Earth and an Object of Mass $m$

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 2
\begin{tikzpicture}

\draw[thick] (0,0) circle (3);

\fill (0,0) circle (0.05);
\node[below left] at (0,0) {$O$};

\draw[dashed] (0,0) -- (2.1,2.1);
\node[above] at (1,1.2) {$R_e$};

\node[above left] at (-2,2.4) {$M_e$};

\fill (2.7,2.7) circle (0.08);
\node[above right] at (2.3,2.8) {$m$};

\draw[<->] (2.1,2.1) -- (2.7,2.7);
\node[right] at (2.4,2.4) {$h$};

\draw[->,thick] (2.7,2.7) -- (3.2,3.2);
\node[right] at (3.8,3.2) {${\bf E}_z$};

\end{tikzpicture}
```

\begin{align}
    {\bf F} = G\frac{M_e m}{\lp R_e+h\rp^2}\lp-{\bf E}_z\rp.
\end{align}

Since $h \ll R_e$, we have $\lp R_e+h\rp^2 \approx R_e^2$, giving:
\begin{align}
    {\bf F} = G\frac{M_e}{R_e^2}m\lp-{\bf E}_z\rp = mg\lp-{\bf E}_z\rp,
\end{align}
where $g = G M_e / R_e^2 = 9.81$ m/s$^2 = 32.2$ ft/s$^2$.

Earth's parameters:

- $M_e = 5.972 \times 10^{24}$ kg
- $R_e = 6.371 \times 10^6$ m

{{< video https://youtu.be/nHnDuCWtacY title="Gravitational force model" >}}

