---
title: "Momenta and Impulses of Particles"
---

## Linear Momentum and Its Conservation

Recall the linear momentum ${\bf G} = m{\bf v}$.

### Linear Impulse and Linear Momentum

The integral form of the balance of linear momentum:
\begin{align}
    {\bf G}(t_1)-{\bf G}(t_0) = \int_{t_0}^{t_1}{\bf F}\,dt.
\end{align}

The time integral of a force is its **linear impulse**. This form is more general than ${\bf F}=m{\bf a}$ because it does not require ${\bf v}$ to be differentiable.

### Conservation of Linear Momentum

::: {.callout-tip title="Think!"}
**Question:** Under what conditions is the linear momentum of a system conserved in a direction ${\bf c}$?

**Answer:**
\begin{align}
    \frac{d}{dt}({\bf G}\cdot{\bf c}) = {\bf F}\cdot{\bf c}+{\bf G}\cdot\dot{\bf c} = 0.
\end{align}
For a constant ${\bf c}$, ${\bf G}\cdot{\bf c}$ is conserved if and only if ${\bf F}\cdot{\bf c} = 0$ — i.e., no force in that direction.
:::

### Example

::: {.callout-warning title="Example"}
**Question:** Is the linear momentum of a projectile conserved? In a certain direction?

**Answer:** With ${\bf W} = -mg{\bf E}_y$, linear momentum is conserved in the ${\bf E}_x$ direction (no force there) but not in ${\bf E}_y$.
:::

## The Moment of a Force

::: {.callout-tip title="Think!"}
**Question:** How do you calculate the moment of a force about a point?

**Answer:** The moment of force ${\bf F}$ applied at $A$ about point $P$ is:
\begin{align}
    {\bf M}^P = \lp{\bf r}_A-{\bf r}_P\rp\times {\bf F}.
\end{align}
:::

::: {.callout-warning title="Example"}
**Question:** Given
\begin{align*}
    {\bf F} &= F_x{\bf E}_x+F_y{\bf E}_y+F_z{\bf E}_z,\\
    {\bf r}_P &= -2{\bf E}_x+3{\bf E}_y+{\bf E}_z,\\
    {\bf r}_A &= -{\bf E}_x+2{\bf E}_y-{\bf E}_z,
\end{align*}
calculate ${\bf M}^O$ and ${\bf M}^P$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick,->] (0,0) -- (0,1) node[above] {${\bf E}_z$};
    \draw[thick,->] (0,0) -- (-1.3,-1) node[left] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (1.3,-1) node[right] {${\bf E}_y$};
    \fill (0,0) circle (0.05);
    \node at (-0.3,0.1) {$O$};

    \fill (4,2) circle (0.05);
    \node at (4,2.5) {$P$};

    \fill (3,-1) circle (0.05);
    \node at (3,-1.5) {$A$};

    \draw[thick,->] (3,-1) -- (6,0) node[midway,above] {${\bf F}$};
    \draw[->] (0,0) -- (3,-1) node[midway,above] {${\bf r}_{A/O}$};
    \draw[->] (0,0) -- (4,2) node[midway,above] {${\bf r}_{P/O}$};
    \draw[->] (4,2) -- (3,-1) node[midway,right] {${\bf r}_{A/P}$};
\end{tikzpicture}
```

**Answer:** ...
:::

## Angular Momentum and Its Conservation

Let ${\bf r}$ be the position vector of a particle relative to a fixed point $O$, and ${\bf v}$ its absolute velocity. The **angular momentum** relative to $O$:
\begin{align}
    {\bf H}^O = {\bf r}\times m{\bf v} = {\bf r}\times{\bf G}.
\end{align}

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \draw[thick,->] (0,0) -- (0,1) node[above] {${\bf E}_z$};
    \draw[thick,->] (0,0) -- (-1.3,-1) node[left] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (1.3,-1) node[right] {${\bf E}_y$};
    \fill (0,0) circle (0.05);
    \node at (-0.3,0.1) {$O$};

    \fill[gray!40] (3,1) circle (0.3);
    \node at (2.8,1.5) {$m$};

    \draw[thick,->] (3,1) -- (6,0) node[right] {${\bf v}$};
    \draw[thick,->] (0,0) -- (3,1) node[midway,above] {${\bf r}$};
    \draw[thick,blue,->] (3,1) -- (4,4) node[above] {${\bf H}^O$};
\end{tikzpicture}
```

### Angular Momentum Theorem

From the BoLM:
\begin{align}
    \dot{\bf H}^O = \frac{d}{dt}\lp{\bf r}\times m{\bf v}\rp = \underbrace{{\bf v}\times m{\bf v}}_{{\bf 0}}+{\bf r}\times m\dot{\bf v} = {\bf r}\times {\bf F}.
\end{align}
So $\dot{\bf H}^O = {\bf M}^O$ (the angular momentum theorem).

### Conservation of Angular Momentum

::: {.callout-tip title="Think!"}
**Question:** Under what conditions is ${\bf H}^O\cdot{\bf c}$ conserved?

**Answer:** For a constant ${\bf c}$: if and only if $\lp{\bf r}\times{\bf F}\rp\cdot{\bf c} = 0$.
:::

### Central Force Problems

A **central force** problem has ${\bf F}$ parallel to ${\bf r}$, so $\dot{\bf H}^O = {\bf 0}$ and ${\bf H}^O = h{\bf h}$ = constant.

The vectors ${\bf r}$ and ${\bf v}$ remain in a fixed plane through $O$. Choosing ${\bf E}_z = {\bf h}$:
\begin{align}
    {\bf H}^O = h{\bf E}_z = {\bf r}(t_0)\times m{\bf v}(t_0).
\end{align}

### Kepler's Problem

The gravitational force on a planet of mass $m$ by the sun of mass $M$:
\begin{align}
    {\bf F} = -\frac{GmM}{\lnorm{\bf r}\rnorm^3}{\bf r}, \qquad U = -\frac{GmM}{\lnorm{\bf r}\rnorm}.
\end{align}

Two conserved quantities:
\begin{align}
    E &= \frac{1}{2}m(\dot{r}^2+r^2\dot{\theta}^2)-\frac{GMm}{r},\\
    h &= {\bf H}^O\cdot{\bf E}_z = mr^2\dot{\theta}.
\end{align}

Watch this [video on Kepler's laws](https://www.youtube.com/watch?v=pdst6HQkdrc).

### Particle on a Smooth Cone

Show that ${\bf H}^O\cdot{\bf E}_z$ is conserved.
