---
title: "Momenta and Impulses of Particles"
---

## Linear Momentum and Its Conservation

Recall the linear momentum ${\bf G} = m{\bf v}$.

### Linear Impulse and Linear Momentum

The integral form of the balance of linear momentum:
\begin{align}
    {\bf G}(t_1)-{\bf G}(t_0) = \int_{t_0}^{t_1}{\bf F}\,dt.
\end{align}

The time integral of a force is its **linear impulse**. This form is more general than ${\bf F}=m{\bf a}$ because it does not require ${\bf v}$ to be differentiable.

{{< video https://youtu.be/-AHQOYy3VBk title="Linear momentum and conservation" >}}

### Conservation of Linear Momentum

::: {.callout-tip title="Think!"}
**Question:** Under what conditions is the linear momentum of a system conserved?

::: {.callout-note title="Answer" collapse="true"}
…
:::
:::

::: {.callout-tip title="Think!"}
**Question:** Under what conditions is the linear momentum of a system conserved in a direction ${\bf c}$?

::: {.callout-note title="Answer" collapse="true"}
Suppose that the component of ${\bf G}$ in the direction of a given vector ${\bf c}$ is conserved:
\begin{align}
    \frac{d}{dt}({\bf G}\cdot{\bf c}) = 0.
\end{align}
This means that
\begin{align}
    \dot{\bf G}\cdot{\bf c}+{\bf G}\cdot\dot{\bf c} = {\bf F}\cdot{\bf c}+{\bf G}\cdot\dot{\bf c} = 0.
\end{align}
Thus, given a vector ${\bf c}$,
$${\bf G}\cdot{\bf c} \text{ is conserved if, and only if, } {\bf F}\cdot{\bf c}+{\bf G}\cdot\dot{\bf c} = 0.$$
If ${\bf c}$ is a constant vector, ${\bf G}\cdot{\bf c}$ is conserved if, and only if, ${\bf F}\cdot{\bf c} = 0$. From the BoLM, this means that there is no force in this constant direction.
:::
:::

{{< video https://youtu.be/_3a0CbiLJ18 title="Conservation of linear momentum along a fixed direction" >}}

### Example

::: {.callout-warning title="Example"}
**Question:** Is the linear momentum of a projectile conserved? In a certain direction?

::: {.callout-note title="Answer" collapse="true"}
With ${\bf W} = -mg{\bf E}_y$, linear momentum is conserved in the ${\bf E}_x$ direction (no force there) but not in ${\bf E}_y$.
:::
:::

## The Moment of a Force

::: {.callout-tip title="Think!"}
**Question:** How do you calculate the moment of a force about a point?

::: {.callout-note title="Answer" collapse="true"}
The moment of force ${\bf F}$ applied at $A$ about point $P$ is:
\begin{align}
    {\bf M}^P = \lp{\bf r}_A-{\bf r}_P\rp\times {\bf F}.
\end{align}
:::
:::

::: {.callout-warning title="Example"}
**Question:** Given
\begin{align*}
    {\bf F} &= F_x{\bf E}_x+F_y{\bf E}_y+F_z{\bf E}_z,\\
    {\bf r}_P &= -2{\bf E}_x+3{\bf E}_y+{\bf E}_z,\\
    {\bf r}_A &= -{\bf E}_x+2{\bf E}_y-{\bf E}_z,
\end{align*}
calculate ${\bf M}^O$ and ${\bf M}^P$.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 4
\begin{tikzpicture}
    \draw[thick,->] (0,0) -- (0,1) node[above] {${\bf E}_z$};
    \draw[thick,->] (0,0) -- (-1.3,-1) node[left] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (1.3,-1) node[right] {${\bf E}_y$};
    \fill (0,0) circle (0.05);
    \node at (-0.3,0.1) {$O$};

    \fill (4,2) circle (0.05);
    \node at (4,2.5) {$P$};

    \fill (3,-1) circle (0.05);
    \node at (3,-1.5) {$A$};

    \draw[thick,->] (3,-1) -- (6,0) node[midway,above] {${\bf F}$};
    \draw[->] (0,0) -- (3,-1) node[midway,above] {${\bf r}_{A/O}$};
    \draw[->] (0,0) -- (4,2) node[midway,above] {${\bf r}_{P/O}$};
    \draw[->] (4,2) -- (3,-1) node[midway,right] {${\bf r}_{A/P}$};
\end{tikzpicture}
```

::: {.callout-note title="Answer" collapse="true"}
...
:::
:::

{{< video https://youtu.be/qiFdls4aFsw title="Review – calculating the moment of a force" >}}

## Angular Momentum and Its Conservation

Let ${\bf r}$ be the position vector of a particle relative to a fixed point $O$, and ${\bf v}$ its absolute velocity. The **angular momentum** relative to $O$:
\begin{align}
    {\bf H}^O = {\bf r}\times m{\bf v} = {\bf r}\times{\bf G}.
\end{align}
In Cartesian coordinates:
\begin{align}
    {\bf H}^O = m\lp y\dot{z}-z\dot{y}\rp{\bf E}_x+m\lp z\dot{x}-x\dot{z}\rp{\bf E}_y+m\lp x\dot{y}-y\dot{x}\rp{\bf E}_z.
\end{align}
In cylindrical-polar coordinates:
\begin{align}
    {\bf H}^O = -mzr\dot{\theta}{\bf e}_r+m\lp z\dot{r}-r\dot{z}\rp{\bf e}_\theta+mr^2\dot{\theta}{\bf E}_z.
\end{align}

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 4
\begin{tikzpicture}
    \draw[thick,->] (0,0) -- (0,1) node[above] {${\bf E}_z$};
    \draw[thick,->] (0,0) -- (-1.3,-1) node[left] {${\bf E}_x$};
    \draw[thick,->] (0,0) -- (1.3,-1) node[right] {${\bf E}_y$};
    \fill (0,0) circle (0.05);
    \node at (-0.3,0.1) {$O$};

    \fill[gray!40] (3,1) circle (0.3);
    \node at (2.8,1.5) {$m$};

    \draw[thick,->] (3,1) -- (6,0) node[right] {${\bf v}$};
    \draw[thick,->] (0,0) -- (3,1) node[midway,above] {${\bf r}$};
    \draw[thick,blue,->] (3,1) -- (4,4) node[above] {${\bf H}^O$};
\end{tikzpicture}
```

{{< video https://youtu.be/RohK7Sk-pCg title="Angular momentum definition" >}}

### Angular Momentum Theorem

From the BoLM:
\begin{align}
    \dot{\bf H}^O = \frac{d}{dt}\lp{\bf r}\times m{\bf v}\rp = \underbrace{{\bf v}\times m{\bf v}}_{{\bf 0}}+{\bf r}\times m\dot{\bf v} = {\bf r}\times {\bf F}.
\end{align}
So $\dot{\bf H}^O = {\bf M}^O$ (the angular momentum theorem).

{{< video https://youtu.be/5oHdJJfxGvk title="Balance of angular momentum" >}}

### Conservation of Angular Momentum

::: {.callout-tip title="Think!"}
**Question:** Under what conditions is the angular momentum of a system conserved?

::: {.callout-note title="Answer" collapse="true"}
…
:::
:::

::: {.callout-tip title="Think!"}
**Question:** Under what conditions is ${\bf H}^O\cdot{\bf c}$ conserved?

::: {.callout-note title="Answer" collapse="true"}
Suppose that the component of ${\bf H}^O$ in the direction of a given vector ${\bf c}$ is conserved:
\begin{align}
    \frac{d}{dt}\lp{\bf H}^O\cdot{\bf c}\rp = 0.
\end{align}
Then,
\begin{align}
    \frac{d}{dt}\lp{\bf H}^O\cdot{\bf c}\rp = \dot{\bf H}^O\cdot{\bf c}+{\bf H}^O\cdot\dot{\bf c} = \lp{\bf r}\times{\bf F}\rp\cdot{\bf c}+{\bf H}^O\cdot\dot{\bf c}.
\end{align}
Consequently, for a given vector ${\bf c}$,
$${\bf H}^O\cdot{\bf c} \text{ is conserved if, and only if, } \lp{\bf r}\times{\bf F}\rp\cdot{\bf c}+{\bf H}^O\cdot\dot{\bf c}=0.$$
If ${\bf c}$ is a constant vector, then ${\bf H}^O\cdot{\bf c}$ is conserved if, and only if, $\lp{\bf r}\times {\bf F}\rp\cdot{\bf c} = 0$.
:::
:::

{{< video https://youtu.be/9TFLI5Tjiv8 title="Conservation of angular momentum" >}}

### Central Force Problems

::: {.callout-warning title="Example"}
**Question:** Consider the motion of the Earth and the Moon. What are the forces acting on the Moon? Are any quantities conserved?

::: {.callout-note title="Answer" collapse="true"}
Yes. Energy and angular momentum are conserved.
:::
:::

A **central force** problem is one where ${\bf F}$ is parallel to ${\bf r}$. The angular momentum theorem implies $\dot{\bf H}^O = {\bf r}\times{\bf F} = {\bf 0}$, so
\begin{align}
    {\bf H}^O = h{\bf h} = \text{constant} = {\bf r}\times m{\bf v},
\end{align}
where $h$ and ${\bf h}$ are constant.

The vectors ${\bf r}$ and ${\bf v}$ form a plane with constant unit normal ${\bf h}$. This plane passes through $O$ and is fixed. Given initial conditions ${\bf r}(t_0)$ and ${\bf v}(t_0)$, we can choose a cylindrical polar coordinate system such that ${\bf E}_z = {\bf h}$, ${\bf r} = r{\bf e}_r$, and ${\bf v} = \dot{r}{\bf e}_r+r\dot{\theta}{\bf e}_\theta$. To do this, it suffices to choose ${\bf E}_z$ so that
\begin{align}
    {\bf H}^O = h{\bf E}_z = {\bf r}(t_0)\times m{\bf v}(t_0).
\end{align}

### Kepler's Problem

The gravitational force on a planet of mass $m$ by the sun of mass $M$ is conservative:
\begin{align}
    \begin{split}
        & {\bf F} = -\frac{GmM}{\lnorm{\bf r}\rnorm^2}\frac{\bf r}{\lnorm{\bf r}\rnorm} = -\frac{\partial U}{\partial {\bf r}},\\
        & U = -\frac{GmM}{\lnorm{\bf r}\rnorm}.
    \end{split}
\end{align}

The BoLM in cylindrical polar coordinates yields:
\begin{align}
    \begin{split}
        & m\ddot{r}-mr\dot{\theta}^2 = -\frac{GMm}{r^2},\\
        & mr\ddot{\theta}+2m\dot{r}\dot{\theta} = 0.
    \end{split}
\end{align}

Two conserved quantities:
\begin{align}
    E &= \frac{1}{2}m(\dot{r}^2+r^2\dot{\theta}^2)-\frac{GMm}{r},\\
    h &= {\bf H}^O\cdot{\bf E}_z = mr^2\dot{\theta}.
\end{align}

Watch this [video on Kepler's laws](https://www.youtube.com/watch?v=pdst6HQkdrc).

### Particle on a Smooth Cone

Show that ${\bf H}^O\cdot{\bf E}_z$ is conserved.


## Summary

**Linear impulse–momentum**: $\int_{t_A}^{t_B}\mathbf{F}\,dt = \mathbf{G}_B - \mathbf{G}_A$.

**Angular momentum** about fixed $O$: $\mathbf{H}^O = \mathbf{r}\times m\mathbf{v}$.

**BoAM**: $\mathbf{M}^O = \dot{\mathbf{H}}^O$.

**Conservation** of $\mathbf{G}$ if $\mathbf{F}=\mathbf{0}$; conservation of $\mathbf{H}^O$ if $\mathbf{M}^O=\mathbf{0}$.

{{< video https://youtu.be/MEwpRoLB7KM title="Conservations – linear vs. angular momentum vs. energy" >}}


## Exercises

*The following problems are from Set 12 – Impulse and Momentum.*

**1.** [MKB 03-149] Straightforward application of the linear impulse–momentum equation. *(ans. $v=1.218$ m/s down)*

**2.** [MKB 03-159] The block is subjected to a time-varying force $P(t)$ shown in the plot; $P=0$ for $t>3$ s. *(ans. $t_s=3.69$ s)*

{{< video https://youtu.be/ugj4fefUN4c title="Solution – Problem 03/159" >}}


**3.** [MKB 03-161] Use Newton's third law. Recall $\int_{x_A}^{x_B}f(x)\,dx = F_{\mathrm{avg}}(x_B-x_A)$. *(ans. $v_f=0.00264$ m/s, $F_{\mathrm{avg}}=59.5$ N)*

{{< video https://youtu.be/Q-F7ew3jFuQ title="Solution – Problem 03/161" >}}


**4.** [03-167] *(ans. $R_x=559$ lb, $R_y=218$ lb)*

**5.** [03-177] Central force problem – what quantities are conserved? *(ans. $v_P=17\,723$ mi/hr)*

**6.** [03-185] *(ans. $v_B=5.43$ m/s)*

{{< video https://youtu.be/V3ussTTlrs8 title="Solution – Problem 03/185" >}}


**7.** [03-181] *(ans. $|H|=389$ N·m·s, $|M|=260$ N·m)*

**8.** [03-192] Label $r$ as $R$ and the requested angle $\theta$ as $\beta$. The particle moves on a surface of revolution $z^2+(r-1.15R)^2=R^2$. Use conservation of total energy and $\mathbf{E}_z$-component of $\mathbf{H}^O$. *(ans. $\theta=52.9^\circ$)*
