---
title: "Lagrange's Equations of Motion for a System of Particles"
---

## Constraints

Put in simple terms, a holonomic constraint is a constraint which can be written in terms of position coordinates, e.g. $x=0$. A nonholonomic constraint is one which cannot be written in terms of position coordinates, e.g. ${\bf v}\cdot{\bf e}_\theta = 0$.

::: {.callout-tip title="Think!"}

Question: Consider a particle constrained to move on a horizontal surface $z=0$. Is this constraint holonomic?

::: {.callout-note title="Answer" collapse="true"}
...
:::

:::

::: {.callout-tip title="Think!"}

Question: Consider two particles connected by a rigid massless rod of length $\ell$. What constraint is this system of two particles subject to? Is this constraint holonomic or nonholonomic?

::: {.callout-note title="Answer" collapse="true"}
...
:::

:::

::: {.callout-tip title="Think!"}

Question: Consider the system in the previous question where $m_2$ is subject to a knife-edge constraint ${\bf v}_2\cdot{\bf e}_2=0$. Is this constraint holonomic or nonholonomic.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 6
\usetikzlibrary{arrows.meta,calc}
\begin{tikzpicture}[
    >=Stealth,
    every node/.style={font=\itshape},
    Evec/.style={-{Stealth}, gray, thick},
    evec/.style={-{Stealth}, gray!80, thick},
    rvec/.style={-{Stealth}, line width=1.1pt, color=blue!40!black}
]
% --- parameters ---
\coordinate (O)  at (6,0);
\def\myang{25}
\def\mylen{9.4}
\coordinate (m1) at (1.2,3.5);
\coordinate (m2) at ($(m1)+(\myang:\mylen)$);
% --- global frame E1, E2 at O ---
\draw[Evec] (O) -- ++(3,0) node[right] {${\bf E}_1$};
\draw[Evec] (O) -- ++(0,2.6) node[above] {${\bf E}_2$};
% --- rod between m1 and m2 ---
\draw[thick] (m1) -- (m2);
\node at ($(m1)!0.5!(m2)+(0,0.35)$) {$\ell$};
% --- reference (dashed) line and angle arc at m1 ---
\draw[dashed] (m1) -- ++(3,0);
\draw (m1)++(0.8,0) arc (0:\myang:0.8);
\node at ($(m1)+(\myang/2:1.05)$) {$\vartheta$};
% --- local frame e1, e2 at m1 ---
\draw[evec] (m1) -- ++(\myang:1.7) node[right] {${\bf e}_1$};
\draw[evec] (m1) -- ++(\myang+90:1.7) node[above] {${\bf e}_2$};
% --- m1: filled disk ---
\filldraw[gray!60, draw=black, thick] (m1) circle (0.55);
\node[below] at ($(m1)+(0,-0.6)$) {$m_1$};
% --- m2: ring/pulley ---
\draw[thick] (m2) circle (0.55);
\draw[thick] ($(m2)+(\myang-15:0.55)$) -- ($(m2)+(\myang+165:0.55)$);
\node[right] at ($(m2)+(0.3,-0.75)$) {$m_2$};
% --- position vectors r1, r2 from O ---
\draw[rvec] (O) -- (m1) node[pos=0.45,below left=1pt] {${\bf r}_1$};
\draw[rvec] (O) -- ($(m2)+(\myang+180:0.55)$) node[pos=0.55,right=2pt] {${\bf r}_2$};
\end{tikzpicture}
```

::: {.callout-note title="Answer" collapse="true"}
...
:::

:::

Frobenius' integrability criteria may be used to determine definitely if a constraint or a system of constraints is holonomic or not. The criterion is beyond the scope of the class. Here, we restrict our attention to holonomic constraints.

## Lagrange's Prescription on Constraints

Consider a constraint of the form
\begin{align*}
    {\bf f}_1\cdot{\bf v}_1+\hdots+{\bf f}_N\cdot{\bf v}_N + e = 0,
\end{align*}
then the associated Lagrange prescription of the constraint force is $\mu{\bf f}_K$ applied at particle $K$, $K=1,\hdots,n$.

::: {.callout-tip title="Think!"}

Question: Prescribe the constraint force for a particle constrained on a plane $z = 0$.

::: {.callout-note title="Answer" collapse="true"}
This constraint may be written as ${\bf v}\cdot{\bf E}_z = 0$ and the constraint force acting on the particle is ${\bf F} = \mu{\bf E}_z.$
:::

:::

::: {.callout-tip title="Think!"}

Question: Prescribe the constraint force acting on a simple pendulum associated with the constraint that the length of the pendulum is constant.

::: {.callout-note title="Answer" collapse="true"}
A simple pendulum has the constraint ${\bf r} = 0$ which may be written as ${\bf v}\cdot{\bf e}_r = 0$, then the constraint force acting on the particle is ${\bf F} = \mu{\bf e}_r$.
:::

:::

## Unconstrained Equations

Consider a system of $n$ particles with $3n$ coordinates. The system is subject to $m$ holonomic constraints, so the number of degrees of freedom of the system is $3n-m$. The generalized coordinates of the system may be listed as
\begin{align}
    q =
    \begin{bmatrix}
        x^1 & y^1 & z^1 & \cdots & x^n & y^n & z^n
    \end{bmatrix}^T
\end{align}
and the corresponding generalized velocities are
\begin{align}
    u =
    \begin{bmatrix}
        v^{x1} & v^{y1} & v^{z1} & \cdots & v^{xn} & v^{yn} & v^{zn}
    \end{bmatrix}^T
\end{align}
We define $u$ so that later we set $u = \dot{q}$.

We arrange the generalized coordinates such that the first $3n-m$ coordinates are the unconstrained degrees of freedom and the last $m$ coordinates are equal to the holonomic constraints, and can be later set to zero.

::: {.callout-tip title="Think!"}

Question: Generalized coordinates for simple harmonic oscillator.

::: {.callout-note title="Answer" collapse="true"}
$q = [x\ y\ z]$, $u = [v_x\ v_y\ v_z]$.
:::

:::

::: {.callout-tip title="Think!"}

Question: Generalized coordinates for simple pendulum.

::: {.callout-note title="Answer" collapse="true"}
$q = [\theta\ \beta\ z]$, $u = [u_\theta\ u_\beta\ u_z]$ where $\beta = r-\ell$.
:::

:::

The kinetic energy of the system is $T$, then Lagrange's equations of motion of the system are
\begin{align}
    \frac{d}{dt}\lp\frac{\partial T}{\partial u^i}\rp-\frac{\partial T}{\partial q^i} = \sum_{K=1}^n{\bf F}^K\cdot \frac{\partial {\bf r}^K}{\partial q^i}
\end{align}
Equivalently, using the Lagrangian $L = T-U$, where $U$ is the potential energy, the equations of motion of the system may be written as
\begin{align}
    \frac{d}{dt}\lp\frac{\partial L}{\partial u^i}\rp-\frac{\partial L}{\partial q^i} = \sum_{K=1}^n{\bf F}_{nonc}^K\cdot \frac{\partial {\bf r}^K}{\partial q^i}
\end{align}
where ${\bf F}^K$ is the sum of external forces acting on particle $K$ while ${\bf F}_{nonc}^K$ is the sum of nonconservative forces acting on $K$.

### Advantage of Lagrange's Equations

If the system is subject to $m$ holonomic constraints, if you choose the last $m$ generalized coordinates to be equal to the constraints, then the first $m-n$ Lagrange's equations of motion will be uncoupled from the holonomic constraint forces, these are the equations of motion. The last $n-m$ equations can be solved for the holonomic constraint forces. If one does not wish to solve for these forces, one can just use the constrained kinetic energy.

## Constrained Equations

Consider a system of $n$ particles with $3n$ degrees of freedom and $m$ holonomic constraints prescribed using Lagrange's equations of motion. The unconstrained degrees of freedom of the system are arranged in an array of generalized coordinates $q$ and generalized velocities $u$, and the constrained kinetic energy of the system is $\tilde{T}$, then Lagrange's equations of motion of the system are
\begin{align}
    \frac{d}{dt}\lp\frac{\partial \tilde{T}}{\partial u^i}\rp-\frac{\partial \tilde{T}}{\partial q^i} = {\bf F}\cdot{\bf a}_i.
\end{align}
Equivalently, using the constrained Lagrangian $\tilde{L} = \tilde{T}-\tilde{U}$, where $\tilde{U}$ is the constrained potential energy, the equations of motion of the system may be written as
\begin{align}
    \frac{d}{dt}\lp\frac{\partial \tilde{L}}{\partial u^i}\rp-\frac{\partial \tilde{L}}{\partial q^i} = {\bf F}_{nconc}\cdot{\bf a}_i.
\end{align}
If the forces are prescribed with Lagrange's prescription, then ${\bf F}\cdot{\bf a}_i$ and ${\bf F}_{nconc}\cdot{\bf a}_i$ reduce to ${\bf F}_{applied}\cdot{\bf a}_i$ and ${\bf F}_{applied, nconc}\cdot{\bf a}_i$ respectively, that is the applied forces and the nonconservative applied forces.

In many cases, the nonconservative applied forces are zero, hence the commonly known homogeneous form of Lagrange's equation
\begin{align}
    \frac{d}{dt}\lp\frac{\partial \tilde{L}}{\partial u^i}\rp-\frac{\partial \tilde{L}}{\partial q^i} = 0.
\end{align}

## Example: Simple Harmonic Oscillator

Consider a simple harmonic oscillator of mass $m$ connected to a linear spring of unstretched length $\ell_0$ and stiffness $k$. The origin is taken at the location of $m$ when the spring is unstretched so that the position vector of $m$ is ${\bf r} = x{\bf E}_x$ and $x$ denotes the stretch of the spring. The velocity of $m$ is ${\bf v} = v{\bf E}_x$. Find the equation of motion of the mass $m$.

Using unconstrained Lagrangian:
\begin{align}
    \begin{split}
        & q^1 = x,\qquad q^2 = y, \qquad q^3 = z,\\
        & u^1 = v_x,\qquad u^2 = v_y, \qquad u^3 = v_z,\\
    \end{split}
\end{align}
Lagrangian
\begin{align}
    L = T-U  = \frac{m}{2}\lp v_x^2+v_y^2+v_z^2\rp-mgz
\end{align}
Constraint forces $\mu_1{\bf E}_y$, $\mu_2{\bf E}_z$.

Equation along $z$,
\begin{align}
    \begin{split}
        \dot{\bf v}_z+mg = \mu_2.
    \end{split}
\end{align}

Using constrained Lagrangian:

The kinetic energy of the spring is $\tilde{T} = \frac{1}{2}v^2$ and potential energy is $\tilde{U} = \frac{1}{2}kx^2$, thus the Lagrangian is $\tilde{L} = \frac{1}{2}v^2-\frac{1}{2}kx^2$. Here, $m$ has one degree of freedom, so $q = [x]$ and $u = [v]$, so the equation of motion of the system is obtained ...

## Example: Simple Pendulum

\begin{align}
    \begin{split}
        {\bf r} &= \ell{\bf e}_x,\\
        {\bf v} &= \ell\dot{\theta}{\bf e}_y
    \end{split}
\end{align}
Gravitational potential energy

## Example: Double Pendulum

\begin{align}
    \begin{split}
        {\bf r}_1 &= \ell_1\,{}_{1}{\bf e}_x,\\
        {\bf v}_1 &= \ell_1\dot{\theta}_1\,{}_{1}{\bf e}_y,\\
        {\bf r}_2 &= {\bf r}_1+\ell_2\,{}_{2}{\bf e}_x,\\
        {\bf v}_2 &= {\bf v}_1+\ell_2\dot{\theta}_2\,{}_{2}{\bf e}_y
    \end{split}
\end{align}
Gravitational potential energy

## Alternative Principles of Dynamics

Source: O'Reilly's textbook on Intermediate Dynamics, second edition, Section 4.12

- Jean Bernoulli's principle of virtual work (1717) - for statics
- D'Alembert's principle (1743) - extension of the principle of virtual work for dynamics
- Gauss' principle of least constraint (1829)
- Hamilton's principle (1835)

### Principle of Virtual Work and D'Alembert's Principle

Consider the principle of virtual work and D'Alembert's Principle applied to a system of particles. We assume that the particles are subject to a single constraint:
\begin{align*}
    {\bf f}_1\cdot{\bf v}_1+\hdots+{\bf f}_N\cdot{\bf v}_N + e = 0.
\end{align*}
The principle of virtual work and D'Alembert's principle collectively state that the motion of the system of particles is such that the following equation is satisfied:
\begin{align}
    \begin{split}
        \lp {\bf F}_{a1}-m_1\ddot{\bf r}_1\rp\cdot{\bf d}_1+\hdots+\lp {\bf F}_{aN}-m_N\ddot{\bf r}_N\rp\cdot{\bf d}_N = 0,
    \end{split}
\end{align}
for all possible choices of the vectors ${\bf d}_1, \hdots, {\bf d}_N$ that satisfy the condition
\begin{align}
    \begin{split}
        {\bf f}_1\cdot{\bf d}_1+\hdots+{\bf f}_N\cdot{\bf d}_N = 0.
    \end{split}
\end{align}
The vectors ${\bf d}_K$ are known as the virtual displacements and are usually denoted by $\delta {\bf r}_K$. The virtual work performed by the applied force ${\bf F}_{ak}$ is defined as ${\bf F}_{ak}\cdot{\bf d}_K$; thus the equation above states that the combined virtual work of the applied forces ${\bf F}_{aK}$ and the inertial forces $-m_K\ddot{\bf r}_K$ is zero.

How to obtain the equations of motion of the system of particles from the equation above?

Introduce a Lagrange multiplier, which we denote by the scalar function $\mu = \mu(t)$, to accommodate the single constraint on the vectors ${\bf d}_K$:
\begin{align}
    \begin{split}
        \sum_{K=1}^N\lp{\bf F}_{aK}-m\ddot{\bf r}_N\rp\cdot{\bf d}_K+\mu\sum_{K=1}^N{\bf f}_K\cdot{\bf d}_K = 0.
    \end{split}
\end{align}
As a consequence of the Lagrange multiplier $\mu$, the vectors ${\bf d}_K$ can be varied independently. For the previous equation to hold for all such displacements, it is necessary and sufficient that
\begin{align}
    \begin{split}
        m_K\ddot{\bf r}_K = {\bf F}_{aK}+\mu{\bf f}_K \qquad (K=1,\hdots,N).
    \end{split}
\end{align}
These equations are none other than the balances of linear momenta for a system of particles subject to the above constraint, for which the constraint forces are prescribed by use of Lagrange's prescription:
\begin{align}
    \begin{split}
        {\bf F}_{cK} = \mu{\bf f}_K\qquad (K=1,\dots,N).
    \end{split}
\end{align}

### Gauss' Principle of Least Action

Suppose we have a system of $N$ particles that are subject to the constraint
\begin{align*}
    {\bf f}_1\cdot{\bf v}_1+\hdots+{\bf f}_N\cdot{\bf v}_N + e = 0.
\end{align*}
and suppose that the constraint forces are prescribed by using Lagrange's prescription ${\bf F}_{cK} = \mu{\bf f}_K$, $(K=1,\hdots,N)$. Then, in any motion of the system that satisfies the constraint, the constraint forces ${\bf F}_{cK} = \mu{\bf f}_K$ are the least needed to ensure that the constraint is satisfied. That is, Lagrange's prescription is in a sense optimal!

### Hamilton's Principle

For ease of exposition, we restrict our discussion to a single particle of mass $m$ and suppose that the coordinates $q = (q^1,q^2,q^2)$ are used to parametrize the position vector. Hamilton's principle states that the motion of the system between a given initial configuration $q(t_0)$ and a given final configuration $q(t_1)$ is such that it extremizes the integral:
\begin{align}
    \begin{split}
        I = \int_{t_0}^{t_1}Ldt
    \end{split}
\end{align}
where the Lagrangian $L = T(q^i,u^i)-U(q^i)$ and $t_0$ and $t_1$ are fixed instances of time.

There are an infinite number of paths $q(t)$ that can connect two possible configurations, and so finding one that extremizes $I$ appears to be a daunting task.

The necessary conditions for the components of $q(t)$ to extremize $I$ are that $q(t)$ satisfy the following differential equations:
\begin{align}
    \frac{d}{dt}\lp\frac{\partial L}{\partial u^i}\rp-\frac{\partial L}{\partial q^i} = 0\qquad (k = 1,2,3).
\end{align}
In the context of extremizing $I$ (calculus of variations), these are known as the Euler-Lagrange equations.

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
#| fig-width: 4.5
\usetikzlibrary{arrows.meta,decorations.markings}
\begin{tikzpicture}[
    >=Stealth,
    midarrow/.style={
        postaction={decorate,
            decoration={markings, mark=at position 0.52 with {\arrow{Stealth}}}}
    }
]

% --- axes ---
\draw (0,0) -- (9.2,0);
\draw (0,0) -- (0,8.2);

% --- tick marks ---
\foreach \x in {0.4,0.8,...,9.0} \draw (\x,0) -- (\x,-0.08);
\foreach \y in {0.4,0.8,...,8.0} \draw (0,\y) -- (-0.08,\y);

% --- endpoints ---
\coordinate (A) at (1,1);   % (t0, q(t0))
\coordinate (B) at (7.4,7.4); % (t1, q(t1))

% --- dashed guide lines ---
\draw[dashed] (0,1) -- (A) -- (1,0);
\draw[dashed] (0,7.4) -- (B) -- (7.4,0);

% --- filled endpoint dots ---
\filldraw (A) circle (2.2pt);
\filldraw (B) circle (2.2pt);

% --- family of varied paths, all sharing endpoints A and B ---
\draw[thick,midarrow] (A) .. controls (2.0,4.6) and (4.2,7.0) .. (B);
\draw[thick,midarrow] (A) .. controls (2.4,3.3) and (5.0,6.0) .. (B);
\draw[thick,midarrow] (A) .. controls (3.0,2.3) and (5.6,4.9) .. (B);
\draw[thick,midarrow] (A) .. controls (3.6,1.7) and (6.2,3.6) .. (B);
\draw[thick,midarrow] (A) .. controls (4.3,1.3) and (6.8,2.4) .. (B);

% --- axis labels ---
\node[left]  at (0,7.4) {$q(t_1)$};
\node[left]  at (0,1)   {$q(t_0)$};
\node[left]  at (0,4.6) {$q$};
\node[below] at (1,0)   {$t_0$};
\node[below] at (7.4,0) {$t_1$};
\node[below] at (4.7,0) {$t$};

\end{tikzpicture}
```

## Exercises

Double pendulum

```{r}
#| engine: tikz
#| echo: false
#| fig-align: center
\begin{tikzpicture}
    \coordinate (O) at (0,0);
    \coordinate (B) at (1,-2);
    \coordinate (C) at (3,-4);
    \draw[thick] (O) -- (B);
    \draw[thick] (B) -- (C);
    \draw[dashed] (O) -- ++(0,-2);
    \draw[dashed] (B) -- ++(0,-2);
\end{tikzpicture}
```
